### 2023-12-05 paper/main.tex
paper init

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2024}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2024}

\begin{document}

\twocolumn[
\icmltitle{Explaining probabilistic prediction over the simplex with Shapley compositions}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Firstname1 Lastname1}{equal,yyy}
\icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
\icmlauthor{Firstname3 Lastname3}{comp}
\icmlauthor{Firstname4 Lastname4}{sch}
\icmlauthor{Firstname5 Lastname5}{yyy}
\icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
\icmlauthor{Firstname7 Lastname7}{comp}
%\icmlauthor{}{sch}
\icmlauthor{Firstname8 Lastname8}{sch}
\icmlauthor{Firstname8 Lastname8}{yyy,comp}
%\icmlauthor{}{sch}
%\icmlauthor{}{sch}
\end{icmlauthorlist}

\icmlaffiliation{yyy}{Department of XXX, University of YYY, Location, Country}
\icmlaffiliation{comp}{Company Name, Location, Country}
\icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}

\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
  The concept of Shapley value has been widely use for measuring the contribution of each features on a machine learning model's prediction. However, this has been designed for one-dimensional function's codomain. For multiclass probabilistic classifier, where the output is a discrete probability distribution over the set of more than two possible classes, the output domain is a multidimensional simplex. In this case, people usually apply the concept of Shapley value on each each output dimension one-by-one, in an implicit one-vs-rest setting ignoring the compositional nature of the output distribution. Using the Aitchison geometry of the simplex, coming from the field of compositional data analysis, this paper present a multidimensional extention of the concept of Shapley value, names Shapley composition, for explaining probabilistic predictions one the simplex in machine learning.
\end{abstract}

\section{Introduction}

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{biblio}
\bibliographystyle{icml2024}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{You \emph{can} have an appendix here.}

You can have as much text here as you want. The main body must be at most $8$ pages long.
For the final version, one more page can be added.
If you want, you can use an appendix like this one.

The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.  Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2023-12-06 paper/main.tex
paper update

@@
 \usepackage{amsthm}
+\usepackage{bm}
 
@@
 \twocolumn[
-\icmltitle{Explaining probabilistic prediction over the simplex with Shapley compositions}
+\icmltitle{Explaining probabilistic prediction on the simplex with Shapley compositions}
 
@@
 \begin{abstract}
-  The concept of Shapley value has been widely use for measuring the contribution of each features on a machine learning model's prediction.
+  The concept of Shapley value has been widely use for measuring the contribution of each feature on a machine learning model's prediction.
 However, this has been designed for one-dimensional function's codomain.
-For multiclass probabilistic classifier, where the output is a discrete probability distribution over the set of more than two possible classes, the output domain is a multidimensional simplex.
-In this case, people usually apply the concept of Shapley value on each each output dimension one-by-one, in an implicit one-vs-rest setting ignoring the compositional nature of the output distribution.
-Using the Aitchison geometry of the simplex, coming from the field of compositional data analysis, this paper present a multidimensional extention of the concept of Shapley value, names Shapley composition, for explaining probabilistic predictions one the simplex in machine learning.
+For multiclass probabilistic classifier, where the output is a discrete probability distribution over the set of more than two possible classes, the output lives on a multidimensional simplex.
+In this case, people have been applying the concept of Shapley value on each output dimension one-by-one, in an implicit one-vs-rest setting, ignoring the compositional nature of the output distribution where the relative information between probabilities matter.
+Using the Aitchison geometry of the simplex, coming from the field of compositional data analysis, this paper present a first initiative for a multidimensional extention of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions on the simplex in machine learning.
 \end{abstract}
@@
 \section{Introduction}
+
+Modern machine learning approaches like the one based on deep learning are often regarded as black-boxes making them not reliable for real-life application where the machine learnbing prediction has to be understood.
+These last years, the number of contribution to make models more explainable has therefore increased in the machine learning literature.
+One way to better understand a prediction would be to measure the contribution of each input features on the computation of the model output.
+The concept of Shapley value is now widely used for this purpose \cite{vstrumbelj2014explaining,datta2016} especially since the release of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap}}.
+The Shapley value came from cooperative game theory...
+
+explain shapley in game theory,
+
+How it is applied to ML,
+
+Limitation,
+
+...
+
+\subsection{Contributions}
+
+...
+
+\section{The Shapley value in machine learning}
+
+In this section, we recall the theoretical formulation of the Shapley value for measuring the contribution of each feature on a machine learning prediction.
+
+Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}$.
+Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{In practice, this is usually unknown but the expectation will be replaced by empirical samplings.}.
+Let $S\subseteq \{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
+$\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.
+
+When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$.
+However, when only $\bm{x}_S$ is given, i.e.
+part of the features, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
+The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
+\begin{equation}
+  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
+\end{equation}
+where $v$ is known as the value function:
+\begin{equation}
+  \begin{aligned}
+    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathbb{R},\\
+    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
+  \end{aligned}
+\end{equation}
+where $2^{\mathcal{X}}$ is the set of all subsets of $S$.
+This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
+The whole contribution of the $i$th feature is computing by averaging this quantity over all possible coalitions as follow:
+\begin{equation}
+  \phi_{f,\bm{x},\text{Pr}}(i) = \frac{1}{d!} \sum_{\pi}c_{f,\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),
+\end{equation}
+where $\pi$ is a permutation of the set $S$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
+For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.
+
+This quantity is known as the Shapley value for the $i$th feature.
+It comes from cooperative game theory and is known to be only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}.
+It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values: $(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i)$, which is known as the \emph{efficiency} property.
+
+Like originally developed in game theory, the Shapley value is designed for one-dimensinal codomain of the function $f$.
+For explaining machine learning model with a multidimensional output like for multiclass classifiers, people have been explaining each output dimension one-by-one ignoring the relative information between them.
+Indeed, for application with more than two classes, the probabilistic output of a classifier lives on a multidimensional simplex.
+The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.
+
+\section{Compositional data}
+
+Compositional data carries relative information.
+Each element of a composition \emph{describes a part of some whole} \cite{pawlowskymodeling} like vectors of proportions, concentrations, and discrete probability distributions.
+A $N$-part composition is a vector of $N$ non-zero positive real numbers that sum to a constant $k$.
+Each element of the vector is a part of the \emph{whole} $k$.
+The sample space of compositional data is known as the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$.
+In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}.
+He defined several operations on the simplex which leads to what is called the \emph{Aitchison geometry of the simplex}.
+
+\subsection{The Aitchison geometry of the simplex}
+John Aitchison defined an internal operation called \emph{perturbation}, an external one called \emph{powering} and an inner product \cite{aitchison2001}:
+\begin{itemize}
+\item a \emph{perturbation}:
+  $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$ seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
+\item a \emph{powering}:
+  $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$ seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
+\item an inner product:\\
+  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$.
+\end{itemize}
+$\mathcal{C}(\cdot)$ is the closure operator.
+Since only the relative information matter, scaling factors are irrelevant and a composition $\bm{x}$ is equivalent to $\lambda \bm{x} = [\lambda x_1,\lambda x_2,\dots\lambda x_N]$ for all $\lambda>0$.
+This equivalence is materialized by the closure operator defined for $k>0$ as: $\mathcal{C}\left(\bm{x} \right) = \left[ \frac{k x_1}{\lVert \bm{x} \rVert_1}, \frac{k x_2}{\lVert \bm{x} \rVert_1} ,\dots \frac{k x_N}{\lVert \bm{x} \rVert_1} \right]^T$, where $\bm{x} \in \mathbb{R}_+^{*N}$ and $\displaystyle \lVert \bm{x} \rVert_1 = \sum_{i=1}^N \lvert x_i \rvert$.%Therefore, any vector of positive real numbers can be projected onto the simplex using the closure.
+
+This give to the simplex a $(N-1)$-dimensional Euclidean vector space structure called \emph{Aitchison geometry of the simplex}.
+In this paper, since we are interested in classifiers' outputs as discrete probability distributions, we restrict ourselves to the \emph{probability simplex} where $k=1$.
+
+\subsection{The isometric log-ratio transformation}
+
+...
+
+\section{Shapley compositions on the simplex}
+
+\section{Explaining probabilistic prediction}
+
+\section{Discussion and conclusion}
 

### 2023-12-07 paper/main.tex
edit paper

@@
 \icmltitlerunning{Submission and Formatting Instructions for ICML 2024}
+
+\DeclareMathOperator{\ilr}{ilr}
+\DeclareMathOperator{\tr}{tr}
 
@@
 \section{The Shapley value in machine learning}
+\label{sec:shapley}
 
@@
 \begin{equation}
+  \label{eq:contrib}
   c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
@@
 \begin{equation}
+  \label{eq:valuefunction}
   \begin{aligned}
@@
 
-...
+An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built.
+The projection of a composition (a discrete probability distribution in our case) into this basis defined an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
+This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry.
+Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition, multiplication by a scalar and inner product.
 
-\section{Shapley compositions on the simplex}
+Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
+The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
+The choice of the basis will be discussed latter in Section \ref{sec:explain}.
+
+\section{Shapley on the simplex}
+
+In Section \ref{sec:shapley} we have briefly presented the standard formulation of Shapley value designed for one-dimensional prediction.
+In this Section, we will see how the Aitchison geometry can be used to extend this concept to the multidimensional simplex for explaning probabilistic predictions.
+
+Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$.
+To properly consider the relative information between the probabilities, The outputs of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex.
+We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
+\begin{equation}
+  \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S),
+\end{equation}
+where $\bm{a}\ominus\bm{b}$ is the perturbation $\bm{a} \oplus \left( (-1)\odot \bm{b}\right)$ which correspond to a substraction between compositions $\bm{a}$ and $\bm{b}$, and where:
+\begin{equation}
+  \label{eq:valuefunctionsimplex}
+  \begin{aligned}
+    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
+    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S],
+  \end{aligned}
+\end{equation}
 
 \section{Explaining probabilistic prediction}
+\label{sec:explain}
+
+\subsection{Implementation and experiment}
 

### 2023-12-07 paper/main.tex
edit paper

@@
 %\usepackage[disable,textsize=tiny]{todonotes}
-\usepackage[textsize=tiny]{todonotes}
+% \usepackage[textsize=tiny]{todonotes}
 
@@
     \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
-    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S],
+    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
   \end{aligned}
 \end{equation}
+The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as \cite{pawlowskymodeling}: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
+where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure.
 
-\section{Explaining probabilistic prediction}
+The Shapley quantity expressing the contribution of the $i$th feature on a prediction can simply be expressed on the simplex as the \emph{Shapley composition} $\bm{\phi}(i)$ given by:
+\begin{equation}
+  \bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i) = \frac{1}{d!} \odot \underset{\pi}{\bigoplus}\bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}).
+\end{equation}
+It can be shown (in Appendix \ref{app:properties}) that the linearity and the efficiency properties naturally hold for the Shapley composition:
+\begin{equation}
+  \begin{aligned}
+    \bm{\phi}_{\alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})}(i) &= \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i),\\
+    \underset{i=1}{\overset{d}\bigoplus} \bm{\phi}_{\bm{f}}(i) &= \bm{f}(\bm{x}) \ominus \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{X})].
+  \end{aligned}
+\end{equation}
+This can be seen as a multidimensional extension of the Shapley value framework on the simplex.
+Here, the Shapley quantity is not a scalar anymore, this is a composition living on the probability simplex.
+In the next section, we will see in more details how this can be used to explain the contribution of the features on a multidimensional probabilistic prediction.
+
+\section{Explaining probabilistic prediction with Shapley compositions}
 \label{sec:explain}
 
-\subsection{Implementation and experiment}
+\subsection{About the implementation and the experiment}
 
@@
 \onecolumn
-\section{You \emph{can} have an appendix here.}
 
-You can have as much text here as you want. The main body must be at most $8$ pages long.
-For the final version, one more page can be added.
-If you want, you can use an appendix like this one.
+\section{Linearity and efficiency of the Shapley composition}
+\label{app:properties}
+In this section, we show the linearity of the Shapley composition with respect to the model prediction, and the efficiency property.
 
-The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.
-Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
+\subsection{Linearity}
+
+The Shapley composition is linear, within the Aitchison geometry of the simplex, with respect to linear combination of models' predictions.
+\begin{proof}
+  Let's consider the linear combination of predictions $\bm{h}(\bm{x}) = \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})$. we want to check if:
+  \begin{equation}
+    \label{eq:linearsimplex}
+    \bm{\phi}_{\bm{h}}(i) = \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i).
+  \end{equation}
+  We have:
+  \begin{equation}
+    \begin{aligned}
+      \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{h}(\bm{x})\mid \bm{x}_S] &= \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [\ilr \left(  \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
+                                                                        &= \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [\alpha \ilr \left( \bm{f}(\bm{x})\right) + \beta \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
+                                                                        &= \ilr^{-1} \left( \alpha \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{f}(\bm{x})\right) \mid \bm{x}_S] + \beta \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
+                                                                        &= \alpha \odot \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{f}(\bm{x})\right) \mid \bm{x}_S] \right) \oplus \beta \odot \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
+                                                                        &= \alpha \odot \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{x})\mid \bm{x}_S] \oplus \beta \odot \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{g}(\bm{x})\mid \bm{x}_S].
+    \end{aligned}
+  \end{equation}
+  Therefore, $\bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_S) = \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S)$, meaning that $\bm{v}$ is linear with respect to the learned function or model.
+The linearity of the contribution $\bm{c}$ naturally follows:
+  \begin{equation}
+    \begin{aligned}
+      &\bm{c}_{\bm{h},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_S),\\
+      &~~~~~~~= \left( \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \right) \ominus \left( \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S) \right),\\
+      &~~~~~~~= \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus\alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \ominus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S),\\
+      &~~~~~~~= \alpha \odot \left( \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \right) \oplus \beta \odot \left( \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S)\right),\\
+      &~~~~~~~= \alpha \odot \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \oplus \beta \odot \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S).
+    \end{aligned}
+  \end{equation}
+  And the linearity of the Shap composition:
+  \begin{equation}
+    \begin{aligned}
+      \bm{\phi}_{\bm{h}}(i) &= \frac{1}{d!}  \underset{\pi}{\bigoplus}\bm{c}_{\bm{h},\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),\\
+                            &= \frac{1}{d!}  \underset{\pi}{\bigoplus}\left( \alpha \odot \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \oplus \beta \odot \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S) \right),\\
+                            &= \alpha \odot \left( \frac{1}{d!}  \underset{\pi}{\bigoplus} \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \right) \oplus \beta \odot \left( \frac{1}{d!} \underset{\pi}{\bigoplus}    \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S) \right),\\
+                            &= \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i).
+    \end{aligned}
+  \end{equation}
+\end{proof}
+
+\subsection{Efficiency}
+The efficiency property naturally holds for Shapley compositions within the Aitchison geometry.
+\begin{proof}
+  \begin{equation}
+    \begin{aligned}
+      \underset{i=1}{\overset{d}\bigoplus} \bm{\phi}_{\bm{f}}(i) & = \underset{i=1}{\overset{d}\bigoplus}\left( \frac{1}{d!} \odot \underset{\pi}{\bigoplus} \bm{c}(i,\pi_{\bm{X}}^{<i})\right),\\
+                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \underset{\pi}{\bigoplus} \left( \bm{v}(\pi_{\bm{X}}^{<i+1}) \ominus \bm{v}(\pi_{\bm{X}}^{<i}) \right) \right),\\
+                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \underbrace{\left(\underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<i+1}) \right)}_{\bm{A}_{i+1}} \ominus \underbrace{\left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<i}) \right)}_{\bm{A}_{i}} \right),\\
+                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \bm{A}_{i+1} \ominus \bm{A}_{i} \right),\\
+                                                                 &=  \frac{1}{d!} \odot \left( \bm{A}_{d+1} \ominus \bm{A}_{1} \right), \text{~since we have a telescoping perturbation,}\\
+                                                                 &=  \frac{1}{d!} \odot \left( \left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<d+1}) \right) \ominus \left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<1}) \right) \right),\\
+                                                                 &=  \frac{1}{d!} \odot \left( \left( \underset{\pi}{\bigoplus} \bm{v}(\bm{X}) \right) \ominus \left( \underset{\pi}{\bigoplus} \bm{v}(\bm{X}_{\emptyset}) \right) \right),\\
+                                                                 &= \bm{v}(\bm{X}) \ominus \bm{v}(\bm{X}_{\emptyset}), \text{~since $d!$ is the number of different permutation,}\\
+                                                                 &= \bm{f}(\bm{x}) \ominus \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{X})].
+    \end{aligned}
+  \end{equation}
+\end{proof}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

### 2023-12-07 paper/main.tex
paper edit

@@
 Like originally developed in game theory, the Shapley value is designed for one-dimensinal codomain of the function $f$.
-For explaining machine learning model with a multidimensional output like for multiclass classifiers, people have been explaining each output dimension one-by-one ignoring the relative information between them.
-Indeed, for application with more than two classes, the probabilistic output of a classifier lives on a multidimensional simplex.
+For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison.
+However, this ignores the relative information between each probability and the compositional nature of probability distributions.
+Indeed, the probabilistic output of a classifier lives on a multidimensional simplex.
 The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.
@@
 
-\subsection{About the implementation and the experiment}
+Given a prediction $\bm{f}(\bm{x})$, the Shapley composition $\bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i)$ describes the contribution of the $i$th feature on the prediction.
+The efficiency property shows how the probability distribution moves from the base value, i.e.
+the expected prediction regardless of the current input, to the prediction $\bm{f}(\bm{x})$.
+In the standard Shapley formulation recalled in Section \ref{sec:shapley}, the prediction is one-dimensional such that the Shapley quantity is a scalar.
+In application where there are more than two possible classes, the prediction is multidimensional such that the Shapley quantity is too.
+Both lives on the same space: the probability simplex.
+In this section, we discuss how the set of Shapley compositions can be analysed to better understand the contribution and influence of each features on the prediction.
+
+\subsection{Angles, norms and projections}
+
+\subsection{Visualization}
+
+\subsubsection{Three classes}
+
+\subsubsection{Four classes}
+
+\subsubsection{More classes}
+
+\subsection{About our plementation and experiment}
 
@@
 
+\newpage
 \subsection{Efficiency}
@@
 \end{proof}
+
+\section{Class compositions}
+
+A $k$-class compositions is defined as an unit norm composition going straight to the direction of the $k$th class.
+This is a discrete probability distribution with maximum probability for the $k$th class
 

### 2023-12-08 paper/main.tex
edit paper

@@
 Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{In practice, this is usually unknown but the expectation will be replaced by empirical samplings.}.
-Let $S\subseteq \{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
+Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
 $\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.
@@
   \begin{aligned}
-    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathbb{R},\\
+    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
     S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
@@
 \end{equation}
-where $2^{\mathcal{X}}$ is the set of all subsets of $S$.
+where $2^{\mathcal{X}}$ is the set of all subsets of $I$.
 This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
@@
 \end{equation}
-where $\pi$ is a permutation of the set $S$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
+where $\pi$ is a permutation of the set $I$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
 For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.
@@
 
-\subsection{About our plementation and experiment}
+\subsubsection{Histograms}
+
+\subsection{About our implementation}
+
+In this work, the estimation algorithm we used to compute the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}.
+Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement.
+Each Shapley compositions are therefore adjusted following a similar method as in the SamplingExplainer of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+See Appendix \ref{app:algo} for more details.
 
@@
 A $k$-class compositions is defined as an unit norm composition going straight to the direction of the $k$th class.
-This is a discrete probability distribution with maximum probability for the $k$th class
+This is a discrete probability distribution with maximum probability for the $k$th class ...
+
+\section{Estimation of the Shapley compositions}
+\label{app:algo}
+In this work, we used an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we present in this section.
+
+Let $d$ be the number of features.
+We want to optimally distribute the $m_{\text{max}}$ drawn samples over the $d$ features.
+Let $\hat{\bm{\phi}_i}$ the estimation of the Shapley composition for the $i$th feature.
+We want to minimize the sum of squared errors or distances $\displaystyle\sum_{i=1}^{d} \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a^2$.
+Since $\hat{\bm{\phi}_i}$ is a (Aitchison) sample mean we have: $\tilde{\hat{\bm{\phi}_i}} \approx \mathcal{N}(\tilde{\bm{\phi}}_i, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ and $\tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i \approx \mathcal{N}(\bm{0}, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ where the tilde refers to the ILR transformation.
+Let $\bm{\Delta_i} = \tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i$ and $Z_i = \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a = \| \tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i \|_2 = \| \bm{\Delta}_i \|_2$.
+The expectation of the sum of squared errors is:
+\begin{equation}
+  \begin{aligned}
+    \mathbb{E}\left[ \sum_{i=1}^{d} Z_i^2 \right] &=   \sum_{i=1}^{d} \mathbb{E}\left[ Z_i^2 \right],\\
+                                                  & =\sum_{i=1}^{d} \mathbb{E}\left[ \sum_{j=1}^{d-1}\Delta_{ij}^2 \right],\\
+                                                  & =\sum_{i=1}^{d} \sum_{j=1}^{d-1} \mathbb{E}\left[\Delta_{ij}^2 \right],\\
+                                                  & =\sum_{i=1}^{d} \sum_{j=1}^{d-1} \frac{1}{m_i}\Sigma_{jj}^{(i)}, \text{ since } \Delta_{ij}\approx \mathcal{Z}(0, \frac{1}{m_i}\Sigma_{jj}^{(i)}),\\
+    &= \sum_{i=1}^d \frac{1}{m_i} \tr \bm{\Sigma}^{(i)}.
+  \end{aligned}
+\end{equation}
+
+When a sample is drawn, the feature for which the sample will be used for improving the Shapley composition estimation is chosen to maximize $\frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1}$.
+Like in \cite{vstrumbelj2014explaining}, this is summarized in Algorithm \ref{alg:2}.
+\begin{algorithm}[tb]
+   \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
+   \label{alg:1}
+\begin{algorithmic}
+   \STATE Initialize $\bm{\phi_i}\leftarrow \ilr^{-1}(\bm{0})$
+   \FOR{$1$ {\bfseries to} $m$}
+   \STATE Randomly select a permutation $\pi$ of the set of indexes $\mathcal{I}$,
+   \STATE Randomly select a sample $\bm{w}\in\mathcal{X}$,
+   \STATE ....
+   \ENDFOR
+\end{algorithmic}
+\end{algorithm}
 

### 2023-12-08 3classes_example.py

@@ -49,7 +49,7 @@ print(sum_shap_base)
 # %%
 #SUMMARIZE WITH NORM, COSINE AND INNER PRODUCTS
 
-(norm_shapley, cos_shap_class, cos_shap_shap) = explainer.summarize()
+(norm_shapley, proj_shap_class, cos_shap_shap) = explainer.summarize()
 
 # %%
 #PLOT the 2D ILR SPACE (CORRESPONDING TO THE CHOSEN ILR COMPONENTS LISTED IN BALANCES).

### 2023-12-08 4classes_example.py

@@ -58,7 +58,7 @@ print(sum_shap_base)
 
 #SUMMARIZE WITH NORM, COSINE AND INNER PRODUCTS
 
-(norm_shapley, cos_shap_class, cos_shap_shap) = explainer.summarize()
+(norm_shapley, proj_shap_class, cos_shap_shap) = explainer.summarize()
 
 # %%
 

### 2023-12-08 more_classes_example.py

@@ -70,7 +70,7 @@ print(sum_shap_base)
 # %%
 #SUMMARIZE WITH NORM, COSINE AND INNER PRODUCTS
 
-(norm_shapley, cos_shap_class, cos_shap_shap) = explainer.summarize()
+(norm_shapley, proj_shap_class, cos_shap_shap) = explainer.summarize()
 
 # %%
 #PLOT the a ILR subspace (corresponding to the chosen ILR components listed in balances).

### 2023-12-08 shapleycomposition.py

@@ -69,7 +69,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
                     x_perturb1[k] = x_sampled[k]
                     x_perturb2[k] = x_sampled[k]
             current_phi = ilr(self.model(x_perturb1.reshape(1,-1)), basis=self.basis) - ilr(self.model(x_perturb2.reshape(1,-1)), basis=self.basis)
-            phi[j] += + current_phi
+            phi[j] += current_phi
             (new_mean, new_M2) = welford_update((mean_phi[j], M2_phi[j]), m[j], (current_phi))
             mean_phi[j] = new_mean
             M2_phi[j]   = new_M2
@@ -104,9 +104,9 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         print()
 
         class_vect  = ilr(self.class_compo)
-        cos_shap_class = np.zeros((self.n_class,self.n_feat))
+        proj_shap_class = np.zeros((self.n_class,self.n_feat))
         cos_shap_shap = np.zeros((self.n_feat,self.n_feat))
-        print("Cosine between the Shapley compositions and the class vectors:")
+        print("Projection of the Shapley compositions on the class vectors:")
         print('\t\t',end='')
         for i in range(self.n_feat):
             print('{:10s}'.format('feat. n.'+str(i+1)),end='\t')
@@ -114,8 +114,8 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         for i in range(self.n_class):
             print('{:10s}'.format("class "+str(i+1)+":"), end='\t')
             for j in range(self.n_feat):
-                cos_shap_class[i,j] = np.dot(class_vect[i,:],self.shapley[j,:])/norm_shapley[j]
-                print('{:3.7f}'.format(round(cos_shap_class[i,j], 7)),end='\t')
+                proj_shap_class[i,j] = np.dot(class_vect[i,:],self.shapley[j,:])
+                print('{:3.7f}'.format(round(proj_shap_class[i,j], 7)),end='\t')
             print()
         print()
 
@@ -132,7 +132,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
                 print('{:3.7f}'.format(round(cos_shap_shap[i,j], 7)),end='\t')
             print()
         
-        return (norm_shapley, cos_shap_class, cos_shap_shap)
+        return (norm_shapley, proj_shap_class, cos_shap_shap)
     
[…]

### 2023-12-11 paper/main.tex
edit paper

@@
 A $k$-class compositions is defined as an unit norm composition going straight to the direction of the $k$th class.
-This is a discrete probability distribution with maximum probability for the $k$th class ...
+This is a discrete probability distribution with maximum probability for the $k$th class ...\\
+...\\
+...
 
@@
    \STATE Randomly select a sample $\bm{w}\in\mathcal{X}$,
-   \STATE ....
+   \STATE Construct two instances:
+   \begin{itemize}
+     \item $\bm{b}_1$: which takes the values from $\bm{x}$ for the $i$th feature and the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise,
+     \item $\bm{b}_2$: which takes the values from $\bm{x}$ the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise.
+     \end{itemize}
+   \STATE $\bm{\phi}_i \leftarrow \bm{\phi}_i \oplus \bm{f}(\bm{b}_1) \ominus \bm{f}(\bm{b}_2) $
    \ENDFOR
+   \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m}$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}[tb]
+   \caption{Adaptation of the Algorithm 2 in \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
+   \label{alg:2}
+   \begin{algorithmic}
+     \STATE Initialization: 
 \end{algorithmic}

### 2023-12-12 paper/main.tex
edit paper

@@
 Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement.
-Each Shapley compositions are therefore adjusted following a similar method as in the SamplingExplainer of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
-See Appendix \ref{app:algo} for more details.
+Each Shapley compositions are therefore adjusted following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+See Appendix \ref{app:algo} and \ref{app:correct} for more details.
 
@@
 
-A $k$-class compositions is defined as an unit norm composition going straight to the direction of the $k$th class.
-This is a discrete probability distribution with maximum probability for the $k$th class ...\\
-...\\
-...
+A $k$-class compositions $\bm{c}^{(k)}$ is defined as an unit norm composition going straight to the direction of the $k$th class.
+This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others.
+The $i$th part of the $\bm{c}^{(k)}$ is:
+\begin{equation}
+    c_i^{(k)} = \left\lbrace
+  \begin{array}{@{}ll@{}}
+    1-(N-1)p, & \text{if}\ i=k \\
+    p, & \text{otherwise,}
+  \end{array}\right.
+\end{equation}
+where $p<\frac{1}{N}$. We want the Aitchison norm of each class composition to be one:
+\begin{equation}
+  \begin{aligned}
+    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 &\iff \sqrt{\frac{1}{2N} \sum_i^N \sum_j^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
+    \text{for clarity, we drop the $(k)$ from the equations,}\\
+    &\iff
+  \end{aligned}
+\end{equation}
 
@@
    \begin{algorithmic}
-     \STATE Initialization: 
+     \STATE Initialization: $m_{i} \leftarrow 0$, $\bm{\phi}_i \leftarrow \bm{0}$, $\forall i \in \{1, \dots d\}$,
+     \WHILE{$\displaystyle \sum_{i=1}^d m_i < m_{\text{max}}$}
+     \IF{$\forall i, m_i \leq m_{\text{min}}$}
+     \STATE $j = \underset{i}{\text{argmax}} \left( \frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1} \right)$,
+     \ELSE
+     \STATE pick a $j$ such that $m_j < m_{\text{min}}$,
+     \ENDIF
+     \STATE $\bm{\phi}_j \leftarrow \bm{\phi}_j +$result of Algorithm \ref{alg:1} for the $j$th feature and $m=1$,
+     \STATE update $\bm{\Sigma}^{(j)}$ using an incremental algorithm,
+     \STATE $m_j \leftarrow m_j+1$
+     \ENDWHILE
+     \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m_i}$, $\forall i \in \{1, \dots d\}$.
 \end{algorithmic}
@@
 
+\section{Correction for the Shapley compositions to be \emph{efficient}}
+\label{app:correct}
+
+In practice, the computation of the Shapley values has an exponential time complexity and we do not have necessarily access to the true distribution of the data.
+The Shapley values are therefore approximated using estimation algorithms like for instance the one presented in the previous appendix.
+However, since the obtained values are approximations, they do not necessarily respect the desired efficiency property.
+This point is often overlooked in the literature.
+In this section we write down an adjustment strategy of the estimated Shapley compositions for them to respect the efficiency property.
+This is a similar strategy as in the sampling approximation of the Shapley values in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+
+Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments).
+Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where is the base $\bm{f}_0$ composition, be the error composition on the pertubation of all Shapley compositions, i.e.
+the error making the efficiency property unfulfilled.
+In order to respect the efficiency property, we want this error to be the neutral element of the perturnation, i.e.
+the ``zero'' in the sense of the Aitchison geometry, i.e the uniform distribution.
+We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.
+with a precision likely to be lower) to move more than the ones with a smaller variance (i.e.
+with a precision likely to be higher).
+
+We therefore weight the $i$th adjustment by a scalar $w_i = w\left(\tr\left(\bm{\Sigma}^{(i)}\right)\right)$, where $w$ is an increasing function, and where $\displaystyle \sum_{i=1}^d w_i= 1$.
+Note that the vector of weight is actually a composition too.
+Similarly to the SHAP toolkit implementation, we choose $w$ as:
+\begin{equation}
+  w_i = w\left(\tr\left(\bm{\Sigma}^{(i)}\right)\right) = \frac{v_i}{\displaystyle 1+\sum_{j=1}^{d}v_j},\text{ where } v_i = \frac{\tr\left(\bm{\Sigma}^{(i)}\right)}{\displaystyle \epsilon \max_j \tr\left(\bm{\Sigma}^{(j)}\right)}.
+\end{equation}
+The $i$th estimated Shapley composition is then asjusted as follow:
+\begin{equation}
+  \hat{\bm{\phi}}_i \leftarrow \hat{\bm{\phi}}_i \oplus \left( w_i \odot \bm{s}_{err}\right).
+\end{equation}
+In this way, when $\epsilon$ goes to zero, the efficiency property is respected for the adjusted Shapley compositions and more weight is given to the adjusments of the Shapley compositions with a higher estimation variance.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

### 2023-12-12 shapleycomposition.py

@@ -44,16 +44,16 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         self.pred       = None
         self.shapley   = None
         
-    def explain_instance(self, x, adjust_sum=True):
+    def explain_instance(self, x, adjust_sum=False):
         pi       = list(range(self.n_feat))
         phi      = np.zeros((self.n_feat,self.n_class-1))
         m        = np.zeros(self.n_feat)
-        var_phi  = np.zeros(self.n_feat)
+        tr_var_phi  = np.zeros(self.n_feat)
         mean_phi = np.zeros((self.n_feat, self.n_class-1))
         M2_phi   = np.zeros((self.n_feat, self.n_class-1, self.n_class-1))
         while m.sum() <= self.m_max:
             if (m < self.m_min).sum() == 0:
-                j = np.argmax((var_phi/m) - (var_phi/(m+1)))
+                j = np.argmax((tr_var_phi/m) - (tr_var_phi/(m+1)))
             else:
                 j = np.argwhere(m<self.m_min)[0][0]
             x_sampled = self.train_data[random.randint(0,self.len_tr_data-1),:] #Gen a random sample from the training set
@@ -74,14 +74,14 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
             mean_phi[j] = new_mean
             M2_phi[j]   = new_M2
             m[j] += 1
-            var_phi[j]  = np.trace(new_M2/m[j])
+            tr_var_phi[j]  = np.trace(new_M2/m[j])
         for i in range(self.n_feat):
             phi[i] = phi[i]/m[i]
 
         self.pred = ilr(self.model(x.reshape(1,-1)), basis=self.basis)
         if adjust_sum:
             #adjust the sum of shapley compositions as in https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py (last visit November 2023)
-            v = 1e6 *var_phi/var_phi.max()
+            v = 1e6 *tr_var_phi/tr_var_phi.max()
             sum_error = self.pred - phi.sum(axis=0) - self.base
             adj = np.ones((self.n_feat,self.n_class))/self.n_class
             for i in range(self.n_feat):
@@ -105,7 +105,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
 
         class_vect  = ilr(self.class_compo)
         proj_shap_class = np.zeros((self.n_class,self.n_feat))
[…]

### 2023-12-13 paper/_region_.tex
edit paper

--- 2023-12-08 paper/main.tex
@@
-%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
+\message{ !name(main.tex)}%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
 
@@
 
-\twocolumn[
-\icmltitle{Explaining probabilistic prediction on the simplex with Shapley compositions}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2024
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Firstname1 Lastname1}{equal,yyy}
-\icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
-\icmlauthor{Firstname3 Lastname3}{comp}
-\icmlauthor{Firstname4 Lastname4}{sch}
-\icmlauthor{Firstname5 Lastname5}{yyy}
-\icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
-\icmlauthor{Firstname7 Lastname7}{comp}
-%\icmlauthor{}{sch}
-\icmlauthor{Firstname8 Lastname8}{sch}
-\icmlauthor{Firstname8 Lastname8}{yyy,comp}
-%\icmlauthor{}{sch}
-%\icmlauthor{}{sch}
-\end{icmlauthorlist}
-
-\icmlaffiliation{yyy}{Department of XXX, University of YYY, Location, Country}
-\icmlaffiliation{comp}{Company Name, Location, Country}
-\icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}
-
-\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
-\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-  The concept of Shapley value has been widely use for measuring the contribution of each feature on a machine learning model's prediction.
-However, this has been designed for one-dimensional function's codomain.
-For multiclass probabilistic classifier, where the output is a discrete probability distribution over the set of more than two possible classes, the output lives on a multidimensional simplex.
-In this case, people have been applying the concept of Shapley value on each output dimension one-by-one, in an implicit one-vs-rest setting, ignoring the compositional nature of the output distribution where the relative information between probabilities matter.
-Using the Aitchison geometry of the simplex, coming from the field of compositional data analysis, this paper present a first initiative for a multidimensional extention of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions on the simplex in machine learning.
-\end{abstract}
-
-\section{Introduction}
-
-Modern machine learning approaches like the one based on deep learning are often regarded as black-boxes making them not reliable for real-life application where the machine learnbing prediction has to be understood.
-These last years, the number of contribution to make models more explainable has therefore increased in the machine learning literature.
-One way to better understand a prediction would be to measure the contribution of each input features on the computation of the model output.
-The concept of Shapley value is now widely used for this purpose \cite{vstrumbelj2014explaining,datta2016} especially since the release of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap}}.
-The Shapley value came from cooperative game theory...
-
-explain shapley in game theory,
-
-How it is applied to ML,
-
-Limitation,
-
-...
-
-\subsection{Contributions}
-
-...
-
-\section{The Shapley value in machine learning}
-\label{sec:shapley}
-
-In this section, we recall the theoretical formulation of the Shapley value for measuring the contribution of each feature on a machine learning prediction.
-
-Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}$.
-Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{In practice, this is usually unknown but the expectation will be replaced by empirical samplings.}.
-Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
-$\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.
-
-When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$.
-However, when only $\bm{x}_S$ is given, i.e.
-part of the features, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
-The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
-\begin{equation}
-  \label{eq:contrib}
-  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
-\end{equation}
-where $v$ is known as the value function:
-\begin{equation}
-  \label{eq:valuefunction}
-  \begin{aligned}
-    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
-    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
-  \end{aligned}
-\end{equation}
-where $2^{\mathcal{X}}$ is the set of all subsets of $I$.
-This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
-The whole contribution of the $i$th feature is computing by averaging this quantity over all possible coalitions as follow:
-\begin{equation}
-  \phi_{f,\bm{x},\text{Pr}}(i) = \frac{1}{d!} \sum_{\pi}c_{f,\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),
-\end{equation}
-where $\pi$ is a permutation of the set $I$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
-For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.
-
-This quantity is known as the Shapley value for the $i$th feature.
-It comes from cooperative game theory and is known to be only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}.
-It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values: $(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i)$, which is known as the \emph{efficiency} property.
-
-Like originally developed in game theory, the Shapley value is designed for one-dimensinal codomain of the function $f$.
-For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison.
-However, this ignores the relative information between each probability and the compositional nature of probability distributions.
-Indeed, the probabilistic output of a classifier lives on a multidimensional simplex.
-The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.
-
-\section{Compositional data}
-
-Compositional data carries relative information.
-Each element of a composition \emph{describes a part of some whole} \cite{pawlowskymodeling} like vectors of proportions, concentrations, and discrete probability distributions.
-A $N$-part composition is a vector of $N$ non-zero positive real numbers that sum to a constant $k$.
-Each element of the vector is a part of the \emph{whole} $k$.
-The sample space of compositional data is known as the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$.
-In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}.
-He defined several operations on the simplex which leads to what is called the \emph{Aitchison geometry of the simplex}.
-
-\subsection{The Aitchison geometry of the simplex}
-John Aitchison defined an internal operation called \emph{perturbation}, an external one called \emph{powering} and an inner product \cite{aitchison2001}:
-\begin{itemize}
-\item a \emph{perturbation}:
-  $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$ seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
-\item a \emph{powering}:
-  $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$ seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
-\item an inner product:\\
-  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$.
-\end{itemize}
-$\mathcal{C}(\cdot)$ is the closure operator.
-Since only the relative information matter, scaling factors are irrelevant and a composition $\bm{x}$ is equivalent to $\lambda \bm{x} = [\lambda x_1,\lambda x_2,\dots\lambda x_N]$ for all $\lambda>0$.
-This equivalence is materialized by the closure operator defined for $k>0$ as: $\mathcal{C}\left(\bm{x} \right) = \left[ \frac{k x_1}{\lVert \bm{x} \rVert_1}, \frac{k x_2}{\lVert \bm{x} \rVert_1} ,\dots \frac{k x_N}{\lVert \bm{x} \rVert_1} \right]^T$, where $\bm{x} \in \mathbb{R}_+^{*N}$ and $\displaystyle \lVert \bm{x} \rVert_1 = \sum_{i=1}^N \lvert x_i \rvert$.%Therefore, any vector of positive real numbers can be projected onto the simplex using the closure.
-
-This give to the simplex a $(N-1)$-dimensional Euclidean vector space structure called \emph{Aitchison geometry of the simplex}.
-In this paper, since we are interested in classifiers' outputs as discrete probability distributions, we restrict ourselves to the \emph{probability simplex} where $k=1$.
-
-\subsection{The isometric log-ratio transformation}
-
-An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built.
-The projection of a composition (a discrete probability distribution in our case) into this basis defined an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
-This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry.
-Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition, multiplication by a scalar and inner product.
-
-Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
-The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
-The choice of the basis will be discussed latter in Section \ref{sec:explain}.
-
-\section{Shapley on the simplex}
-
-In Section \ref{sec:shapley} we have briefly presented the standard formulation of Shapley value designed for one-dimensional prediction.
-In this Section, we will see how the Aitchison geometry can be used to extend this concept to the multidimensional simplex for explaning probabilistic predictions.
-
-Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$.
-To properly consider the relative information between the probabilities, The outputs of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex.
-We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
-\begin{equation}
-  \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S),
-\end{equation}
-where $\bm{a}\ominus\bm{b}$ is the perturbation $\bm{a} \oplus \left( (-1)\odot \bm{b}\right)$ which correspond to a substraction between compositions $\bm{a}$ and $\bm{b}$, and where:
-\begin{equation}
-  \label{eq:valuefunctionsimplex}
-  \begin{aligned}
-    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
-    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
-  \end{aligned}
-\end{equation}
-The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as \cite{pawlowskymodeling}: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
-where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure.
-
-The Shapley quantity expressing the contribution of the $i$th feature on a prediction can simply be expressed on the simplex as the \emph{Shapley composition} $\bm{\phi}(i)$ given by:
-\begin{equation}
-  \bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i) = \frac{1}{d!} \odot \underset{\pi}{\bigoplus}\bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}).
-\end{equation}
-It can be shown (in Appendix \ref{app:properties}) that the linearity and the efficiency properties naturally hold for the Shapley composition:
-\begin{equation}
-  \begin{aligned}
-    \bm{\phi}_{\alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})}(i) &= \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i),\\
-    \underset{i=1}{\overset{d}\bigoplus} \bm{\phi}_{\bm{f}}(i) &= \bm{f}(\bm{x}) \ominus \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{X})].
-  \end{aligned}
-\end{equation}
-This can be seen as a multidimensional extension of the Shapley value framework on the simplex.
-Here, the Shapley quantity is not a scalar anymore, this is a composition living on the probability simplex.
-In the next section, we will see in more details how this can be used to explain the contribution of the features on a multidimensional probabilistic prediction.
-
+\message{ !name(main.tex) !offset(167) }
 \section{Explaining probabilistic prediction with Shapley compositions}
@@
 
-\subsection{Angles, norms and projections}
+\subsection{Visualization}
 
-\subsection{Visualization}
+The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}.
+This space has the advantage of being intuitive since it is the standard real $(N-1)$-dimensional vector space we are used too.
 
@@
 
+In the three classes case, the space is $2$-dimensional an can simply be visualized in a $2$-dimensional plan plot.
+We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width.
+The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to.
+\begin{figure}
+  \centering
+    \includegraphics[width=\linewidth]{figures/3classes/ilrplot.png}
+  \caption{Shapley compositions in the ILR space for an Iris instance}
+  \label{fig:3classes}
+\end{figure}
+
 \subsubsection{Four classes}
 
+...
 \subsubsection{More classes}
+\label{sec:more}
+...
 
-\subsubsection{Histograms}
+\subsection{Angles, norms and projections}
 
+Some may find the fine analysis of the features contributions in cases with more than four classes, like discussed in Section \ref{sec:more}, tricky.
+Indeed, this require a careful choose of the sequential binary partition (i.e.
+the basis) and of the subspace to visualize.
+However, the Shapley explaination can be summarized by sets of angles, norms and projections.
+
+Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.
+
+Orthogonality can also be investigated ...
+...
+
+\subsection{Histograms}
+
+...
 \subsection{About our implementation}
@@
 Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement.
-Each Shapley compositions are therefore adjusted following a similar method as in the SamplingExplainer of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
-See Appendix \ref{app:algo} for more details.
+Each Shapley compositions are therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+See Appendix \ref{app:algo} and \ref{app:correct} for more details.
 
-\section{Discussion and conclusion}
-
-% In the unusual situation where you want a paper to appear in the
-% references without citing it in the main text, use \nocite
-\nocite{langley00}
-
-\bibliography{biblio}
-\bibliographystyle{icml2024}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
-\onecolumn
-
-\section{Linearity and efficiency of the Shapley composition}
-\label{app:properties}
-In this section, we show the linearity of the Shapley composition with respect to the model prediction, and the efficiency property.
-
-\subsection{Linearity}
-
-The Shapley composition is linear, within the Aitchison geometry of the simplex, with respect to linear combination of models' predictions.
-\begin{proof}
-  Let's consider the linear combination of predictions $\bm{h}(\bm{x}) = \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})$. we want to check if:
-  \begin{equation}
-    \label{eq:linearsimplex}
-    \bm{\phi}_{\bm{h}}(i) = \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i).
-  \end{equation}
-  We have:
-  \begin{equation}
-    \begin{aligned}
-      \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{h}(\bm{x})\mid \bm{x}_S] &= \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [\ilr \left(  \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
-                                                                        &= \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [\alpha \ilr \left( \bm{f}(\bm{x})\right) + \beta \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
-                                                                        &= \ilr^{-1} \left( \alpha \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{f}(\bm{x})\right) \mid \bm{x}_S] + \beta \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
-                                                                        &= \alpha \odot \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{f}(\bm{x})\right) \mid \bm{x}_S] \right) \oplus \beta \odot \ilr^{-1} \left( \mathbb{E}_{\text{Pr}} [ \ilr \left( \bm{g}(\bm{x}) \right) \mid \bm{x}_S] \right),\\
-                                                                        &= \alpha \odot \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{x})\mid \bm{x}_S] \oplus \beta \odot \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{g}(\bm{x})\mid \bm{x}_S].
-    \end{aligned}
-  \end{equation}
-  Therefore, $\bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_S) = \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S)$, meaning that $\bm{v}$ is linear with respect to the learned function or model.
-The linearity of the contribution $\bm{c}$ naturally follows:
-  \begin{equation}
-    \begin{aligned}
-      &\bm{c}_{\bm{h},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{h},\bm{x},\text{Pr}}(\bm{X}_S),\\
-      &~~~~~~~= \left( \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \right) \ominus \left( \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S) \right),\\
-      &~~~~~~~= \alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \oplus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus\alpha \odot \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \ominus \beta \odot \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S),\\
-      &~~~~~~~= \alpha \odot \left( \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S) \right) \oplus \beta \odot \left( \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{v}_{\bm{g},\bm{x},\text{Pr}}(\bm{X}_S)\right),\\
-      &~~~~~~~= \alpha \odot \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \oplus \beta \odot \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S).
-    \end{aligned}
-  \end{equation}
-  And the linearity of the Shap composition:
-  \begin{equation}
-    \begin{aligned}
-      \bm{\phi}_{\bm{h}}(i) &= \frac{1}{d!}  \underset{\pi}{\bigoplus}\bm{c}_{\bm{h},\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),\\
-                            &= \frac{1}{d!}  \underset{\pi}{\bigoplus}\left( \alpha \odot \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \oplus \beta \odot \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S) \right),\\
-                            &= \alpha \odot \left( \frac{1}{d!}  \underset{\pi}{\bigoplus} \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) \right) \oplus \beta \odot \left( \frac{1}{d!} \underset{\pi}{\bigoplus}    \bm{c}_{\bm{g},\bm{x},\text{Pr}}(i,\bm{X}_S) \right),\\
-                            &= \alpha \odot \bm{\phi}_{\bm{f}}(i) \oplus \beta\odot \bm{\phi}_{\bm{g}}(i).
-    \end{aligned}
-  \end{equation}
-\end{proof}
-
-\newpage
-\subsection{Efficiency}
-The efficiency property naturally holds for Shapley compositions within the Aitchison geometry.
-\begin{proof}
-  \begin{equation}
-    \begin{aligned}
-      \underset{i=1}{\overset{d}\bigoplus} \bm{\phi}_{\bm{f}}(i) & = \underset{i=1}{\overset{d}\bigoplus}\left( \frac{1}{d!} \odot \underset{\pi}{\bigoplus} \bm{c}(i,\pi_{\bm{X}}^{<i})\right),\\
-                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \underset{\pi}{\bigoplus} \left( \bm{v}(\pi_{\bm{X}}^{<i+1}) \ominus \bm{v}(\pi_{\bm{X}}^{<i}) \right) \right),\\
-                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \underbrace{\left(\underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<i+1}) \right)}_{\bm{A}_{i+1}} \ominus \underbrace{\left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<i}) \right)}_{\bm{A}_{i}} \right),\\
-                                                                 &=  \frac{1}{d!} \odot \underset{i=1}{\overset{d}\bigoplus}\left( \bm{A}_{i+1} \ominus \bm{A}_{i} \right),\\
-                                                                 &=  \frac{1}{d!} \odot \left( \bm{A}_{d+1} \ominus \bm{A}_{1} \right), \text{~since we have a telescoping perturbation,}\\
-                                                                 &=  \frac{1}{d!} \odot \left( \left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<d+1}) \right) \ominus \left( \underset{\pi}{\bigoplus} \bm{v}(\pi_{\bm{X}}^{<1}) \right) \right),\\
-                                                                 &=  \frac{1}{d!} \odot \left( \left( \underset{\pi}{\bigoplus} \bm{v}(\bm{X}) \right) \ominus \left( \underset{\pi}{\bigoplus} \bm{v}(\bm{X}_{\emptyset}) \right) \right),\\
-                                                                 &= \bm{v}(\bm{X}) \ominus \bm{v}(\bm{X}_{\emptyset}), \text{~since $d!$ is the number of different permutation,}\\
-                                                                 &= \bm{f}(\bm{x}) \ominus \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{X})].
-    \end{aligned}
-  \end{equation}
-\end{proof}
-
-\section{Class compositions}
-
-A $k$-class compositions is defined as an unit norm composition going straight to the direction of the $k$th class.
-This is a discrete probability distribution with maximum probability for the $k$th class ...
-
-\section{Estimation of the Shapley compositions}
-\label{app:algo}
-In this work, we used an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we present in this section.
-
-Let $d$ be the number of features.
-We want to optimally distribute the $m_{\text{max}}$ drawn samples over the $d$ features.
-Let $\hat{\bm{\phi}_i}$ the estimation of the Shapley composition for the $i$th feature.
-We want to minimize the sum of squared errors or distances $\displaystyle\sum_{i=1}^{d} \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a^2$.
-Since $\hat{\bm{\phi}_i}$ is a (Aitchison) sample mean we have: $\tilde{\hat{\bm{\phi}_i}} \approx \mathcal{N}(\tilde{\bm{\phi}}_i, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ and $\tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i \approx \mathcal{N}(\bm{0}, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ where the tilde refers to the ILR transformation.
-Let $\bm{\Delta_i} = \tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i$ and $Z_i = \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a = \| \tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i \|_2 = \| \bm{\Delta}_i \|_2$.
-The expectation of the sum of squared errors is:
-\begin{equation}
-  \begin{aligned}
-    \mathbb{E}\left[ \sum_{i=1}^{d} Z_i^2 \right] &=   \sum_{i=1}^{d} \mathbb{E}\left[ Z_i^2 \right],\\
-                                                  & =\sum_{i=1}^{d} \mathbb{E}\left[ \sum_{j=1}^{d-1}\Delta_{ij}^2 \right],\\
-                                                  & =\sum_{i=1}^{d} \sum_{j=1}^{d-1} \mathbb{E}\left[\Delta_{ij}^2 \right],\\
-                                                  & =\sum_{i=1}^{d} \sum_{j=1}^{d-1} \frac{1}{m_i}\Sigma_{jj}^{(i)}, \text{ since } \Delta_{ij}\approx \mathcal{Z}(0, \frac{1}{m_i}\Sigma_{jj}^{(i)}),\\
-    &= \sum_{i=1}^d \frac{1}{m_i} \tr \bm{\Sigma}^{(i)}.
-  \end{aligned}
-\end{equation}
-
-When a sample is drawn, the feature for which the sample will be used for improving the Shapley composition estimation is chosen to maximize $\frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1}$.
-Like in \cite{vstrumbelj2014explaining}, this is summarized in Algorithm \ref{alg:2}.
-\begin{algorithm}[tb]
-   \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
-   \label{alg:1}
-\begin{algorithmic}
-   \STATE Initialize $\bm{\phi_i}\leftarrow \ilr^{-1}(\bm{0})$
-   \FOR{$1$ {\bfseries to} $m$}
-   \STATE Randomly select a permutation $\pi$ of the set of indexes $\mathcal{I}$,
-   \STATE Randomly select a sample $\bm{w}\in\mathcal{X}$,
-   \STATE ....
-   \ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\message{ !name(main.tex) !offset(360) }
 

### 2023-12-13 paper/main.tex
edit paper

@@
 \subsection{The isometric log-ratio transformation}
+\label{sec:ilr}
 
@@
 
+In the three classes case, the space is $2$-dimensional.
+We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width.
+The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to.
+In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier.
+Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction.
+Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region.
+Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class.
+The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class.
+The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
+\begin{figure}
+  \centering
+  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
+    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplot.png}
+    }
+  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
+    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplotsum.png}
+  }
+  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
+  \label{fig:3classes}
+\end{figure}
+
 \subsubsection{Four classes}
 
+In a four classes examples, the simplex is $3$-dimensional.
+We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}.
+It consists of classifying an $8\times8$ image as representing one of the digit amoung: 0, 1, 2 and 3.
+Since they are 64 pixels as features which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for clarity.
+We again use a SVM classifier with an rbf kernel.
+The same explaintion analysis as before can be applied here but in a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}.
+To better understand of this space is divided in four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule.
+The hydrogens correspond to the vextex and the carbon to the center of the tetrahedron i.e.
+a $3$-dimensional simplex.
+The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
+  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task.
+The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
+  \label{fig:4classesshapsum}
+\end{figure}
+
 \subsubsection{More classes}
+\label{sec:more}
+...
 
-\subsubsection{Histograms}
+\subsection{Summarize explaination with angles, norms and projections}
+
+Some may find the fine analysis of the features contributions in cases with more than four classes, like discussed in Section \ref{sec:more}, tricky.
+Indeed, this require a careful choose of the sequential binary partition (i.e.
+the basis) and of the subspace to visualize.
+However, the Shapley explaination can be summarized by sets of angles, norms and projections.
+
+Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.
+
+Orthogonality can also be investigated ...
+...
+
+\subsection{Histograms}
+
+If one found hard to visualize the proposed Shapley explanation in the ILR space, the Shapley composition can be visualized as histograms like discrete probability distributions.
 
@@
 Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement.
-Each Shapley compositions are therefore adjusted following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+Each Shapley compositions are therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
 See Appendix \ref{app:algo} and \ref{app:correct} for more details.
@@
 \section{Discussion and conclusion}
+
+Compare with standard Shapley ??
+
+We know small expé..., sounds tricky,... first step for a theoretically founded multiclass problems explanations...
+...
 
@@
 
+\newpage
 \section{Class compositions}
 
-A $k$-class compositions $\bm{c}^{(k)}$ is defined as an unit norm composition going straight to the direction of the $k$th class.
+A $k$-class compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
 This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others.
@@
   \begin{aligned}
-    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 &\iff \sqrt{\frac{1}{2N} \sum_i^N \sum_j^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
-    \text{for clarity, we drop the $(k)$ from the equations,}\\
-    &\iff
+    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
+    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
+    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
+    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
+    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
+    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
+    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
   \end{aligned}
 \end{equation}
+To summarize, the $i$th part of a $k$-class compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
+\begin{equation}
+    c_i^{(k)} =
+  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
+     1, & \text{if}\ i=k \\
+     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
+  \end{array}\right.\right).
+\end{equation}
+In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.
 
+\newpage
 \section{Estimation of the Shapley compositions}
@@
 Like in \cite{vstrumbelj2014explaining}, this is summarized in Algorithm \ref{alg:2}.
-\begin{algorithm}[tb]
+\begin{algorithm}
    \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
@@
 
-\begin{algorithm}[tb]
+\begin{algorithm}
    \caption{Adaptation of the Algorithm 2 in \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
@@
 
-\section{Correction for the Shapley compositions to be \emph{efficient}}
+\newpage
+\section{Adjustement of the estimated Shapley compositions for efficiency}
 \label{app:correct}

### 2023-12-13 paper/main.tex
edit paper

@@
 The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
-The choice of the basis will be discussed latter in Section \ref{sec:explain}.
+The choice of the basis will be discussed in Section \ref{sec:moreclasses}.
 
@@
   \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
-    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplot.png}
+    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
     }
   \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
-    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplotsum.png}
+    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
   }
@@
 
-In a four classes examples, the simplex is $3$-dimensional.
+In a four classes example, the simplex is $3$-dimensional.
 We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}.
-It consists of classifying an $8\times8$ image as representing one of the digit amoung: 0, 1, 2 and 3.
-Since they are 64 pixels as features which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for clarity.
+It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3.
+Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity.
 We again use a SVM classifier with an rbf kernel.
-The same explaintion analysis as before can be applied here but in a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}.
-To better understand of this space is divided in four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule.
-The hydrogens correspond to the vextex and the carbon to the center of the tetrahedron i.e.
+The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}.
+To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule.
+The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e.
 a $3$-dimensional simplex.
@@
 \subsubsection{More classes}
-\label{sec:more}
-...
+\label{sec:moreclasses}
+When there are more classes, all the dimension of the ILR space cannot be visualized at once but $2$ or $3$-dimensional subspaces can still be visualized.
+However, in order to choose which ILR components to visualize, one needs to understand what they refer too.
+The next section provides some intuition about that in the context of the explanation of a classifier prediction with more than 4 classes.
+
+\subsection{Groups of part, balances and sequential binary partition}
+
+....
 
@@
 
-Some may find the fine analysis of the features contributions in cases with more than four classes, like discussed in Section \ref{sec:more}, tricky.
-Indeed, this require a careful choose of the sequential binary partition (i.e.
-the basis) and of the subspace to visualize.
-However, the Shapley explaination can be summarized by sets of angles, norms and projections.
-
+Some may find the fine analysis of the features contributions in cases with more than four classes tricky.
+Indeed, this require a careful choose of the sequential binary partition and of the ILR composition to visualize.
+However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections.
 Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.
-
-Orthogonality can also be investigated ...
-...
+The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory.
+The projection of a Shapley composition on the set of class composition informs in favor of, or against which, classes a feature is contribution.
+Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation providing the norm of Shapley compositions, angles between them and their projection on class compositions.
 
@@
 \newpage
+\section{Summarizing the explanation with norms, angles and projections of Shapley compositions}
+
+TODO...
+
+\newpage
 \section{Estimation of the Shapley compositions}

### 2023-12-14 paper/main.tex
edit paper, names in summa

@@
 Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region.
-Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class.
+Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class.
 The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class.
@@
 a $3$-dimensional simplex.
-The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
+The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
 \begin{figure}
@@
 
-\subsection{Summarize explaination with angles, norms and projections}
+\subsection{Angles, norms and projections}
 
 Some may find the fine analysis of the features contributions in cases with more than four classes tricky.
-Indeed, this require a careful choose of the sequential binary partition and of the ILR composition to visualize.
+Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition.
 However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections.
-Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.
+Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction.
 The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory.
-The projection of a Shapley composition on the set of class composition informs in favor of, or against which, classes a feature is contribution.
-Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation providing the norm of Shapley compositions, angles between them and their projection on class compositions.
+The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing.
+Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.
 
@@
 ...
+Features INDEPENDENCE!!
+
+Axiomatic formulation
 
@@
 \newpage
-\section{Class compositions}
+\section{Class-compositions}
 
-A $k$-class compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
+A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
 This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others.
@@
 \end{equation}
-where $p<\frac{1}{N}$. We want the Aitchison norm of each class composition to be one:
+where $p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
 \begin{equation}
@@
 \end{equation}
-To summarize, the $i$th part of a $k$-class compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
+To summarize, the $i$th part of a $k$-class-compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
 \begin{equation}
@@
 \section{Summarizing the explanation with norms, angles and projections of Shapley compositions}
+\label{app:summarize}
 
-TODO...
+The example of Figure \ref{fig:3classes} can be summarized by computing the norms of the Shapley compositions, angles between them and their projection on the set of class-compositions.
 

### 2023-12-14 3classes_example.py

@@ -12,9 +12,9 @@ import plotly.graph_objects as go
 K = 10         #index of the instance you want to test in the test set
 N_class = 3    #number of class, the dimension of the simplex is therefore N_class-1
 
-
 #load the dataset and split into a training and testing set
-X, Y = datasets.load_iris(return_X_y=True)
+data = datasets.load_iris()
+(X, Y) = (data.data, data.target)
 X_train,X_test,Y_train,Y_test = train_test_split(X, Y, test_size=0.2, random_state=0)
 
 #train an SVM classifier
@@ -56,6 +56,8 @@ print(sum_shap_base)
 #plot range [-lim, lim]
 #If shapley_sum is True, the sum of the shapley vectors are summed from the base distribution to the prediction
 
-fig = explainer.plot_ilr_space(balances=[1, 2], lim=3, figsize=750)
+fig = explainer.plot_ilr_space(balances=[1, 2], lim=3, figsize=450,
+                               names_classes=data.target_names, names_features=['sep. length','sep. width','pet. length','pet. width'])
 
-fig = explainer.plot_ilr_space(balances=[1, 2], shapley_sum=True, lim=3, figsize=750)
+fig = explainer.plot_ilr_space(balances=[1, 2], shapley_sum=True, lim=3, figsize=450, 
+                               names_classes=data.target_names, names_features=['sep. length','sep. width','pet. length','pet. width'])

### 2023-12-14 shapleycomposition.py

@@ -27,7 +27,7 @@ class ShapleyExplainer():
     """ Compute the Shapley for each feature with the sampling approximation (Algo 2) in "Explaining prediction models and individual predictions
 with feature contributions" by Erik Štrumbelj and Igor Kononenko.
     """    
-    def __init__(self, model, train_data, n_class, m_min=100, m_max=10000, sbpmatrix=None):
+    def __init__(self, model, train_data, n_class, m_min=100, m_max=10000, sbpmatrix=None, names_classes=None, names_features=None):
         self.model       = model
         self.train_data  = train_data
         self.len_tr_data = train_data.shape[0]
@@ -40,11 +40,32 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
             self.basis = None
         else:
             self.basis = np.flip(sbp_basis(sbpmatrix), axis=0)
-        self.base       = ilr(model(train_data), basis=self.basis).mean(axis=0)
-        self.pred       = None
+        self.base      = ilr(model(train_data), basis=self.basis).mean(axis=0)
+        self.pred      = None
         self.shapley   = None
-        
-    def explain_instance(self, x, adjust_sum=False):
+
+        #Names of the classes and features if not precised in the calling of the plotting function
+        if names_classes is None:
+            self.names_classes = ['class '+str(i+1) for i in range(self.n_class)]
+        elif (type(names_classes) is list) or (type(names_classes) is np.ndarray):
+            if len(names_classes) != self.n_class:
+                raise NameError('The number of strings in the list (or array) of class names must be the number of classes: '+str(self.n_class))
+            else:
+                self.names_classes = names_classes
+        else:
+            raise NameError('names_classes must be a list (or array) of '+str(self.n_class)+' strings.')
+                
+        if names_features is None:
+            self.names_features = ['feature n.'+str(i+1) for i in range(self.n_feat)]
+        elif (type(names_features) is list) or (type(names_features) is np.ndarray):
+            if len(names_features) != self.n_feat:
+                raise NameError('The number of strings in the list (or array) of feature names must be the number of features: '+str(self.n_feat))
+            else:
[…]

### 2023-12-15 paper/main.tex
edit app summa

@@
 \usepackage{hyperref}
+\usepackage{multirow}
 
@@
 Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction.
+It gives the overall contribution of the feature on the prediction, regardless of its direction.
 The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory.
@@
 
-The example of Figure \ref{fig:3classes} can be summarized by computing the norms of the Shapley compositions, angles between them and their projection on the set of class-compositions.
+The Table \ref{tab:normiris} gives, for the Iris classification example of Figure \ref{fig:3classes}, the norm of each Shapley compositions, the cosine similarity between them and their projection on the set of class-compositions.
+Having the highest norm, this confirms that the petal length is the feature that contributed the most on the prediction.
+Having a close to zero inner product with the virginica class-composition, shows that this feature did not contribute to the value of the probability of of this class.
+It contributes neither in favor, nor in the reject of this class.
+Having a positive inner product with the versicolor class-composition and a negative one with the setosa class-composition, this suggest that this feature is going in favor of the class versicolor and against setosa.
+\begin{table}
+  \centering
+  \caption{Norm of the Shapley compositions, projection on the class-compositions and cosine similarity betweeen the Shapley compositions for the Iris classification example.}
+  \begin{tabular}{c|c|ccc|cccc|}
+    \cline{2-9}
+    & \multirow{2}{*}{norm} & \multicolumn{3}{c|}{projection on the class-compositions} & \multicolumn{4}{c|}{cosine similarity between each Shapley composition} \\
+    \cline{3-9}
+     & & setosa & versicolor & virginica & pet. length & pet. width & sep. length & sep. width \\
+    \hline
+    pet. length & 2.20  & -1.91 & 1.90 & 0.02 & 1 &$\cdot$ &$\cdot$ &$\cdot$ \\
+    pet. width & 0.51 & -0.51 & 0.29 & 0.22 & 0.90 & 1 &$\cdot$ &$\cdot$ \\
+    sep. length & 0.13 & -0.08 & 0.13 & -0.05 & 0.93 & 0.69 & 1 &$\cdot$ \\
+    sep. width & 0.11 & -0.11 & 0.04 & 0.07 & 0.77 & 0.97 & 0.49 &1 \\
+    \hline
+  \end{tabular}
+  \label{tab:normiris}
+\end{table}
+
+Other examples ... !!!
 

### 2023-12-15 paper/main.tex
edit paper and add histograms

@@
 \subsection{Groups of part, balances and sequential binary partition}
-
-....
+\ref{sec:balances}
+..TODO..
 
@@
 If one found hard to visualize the proposed Shapley explanation in the ILR space, the Shapley composition can be visualized as histograms like discrete probability distributions.
+Figure \ref{fig:histiris} shows the Shapley compositions of the Iris classification example.
+The more uniform the histogram is, like for the sepal length and width, the less the contribution of the feature is.
+In opposite, the histogram for the petal length as a high value for the versicolor class, relatively to the others, confirming the contribution of the feature toward this class.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/3classes/histo}
+  \caption{Shapley compositions visualized as histograms for the Iris classification example.}
+  \label{fig:histiris}
+\end{figure}
+As another illustration, Figure \ref{fig:histmore} shows the Shapley compositions of the seven classes digit recognition example.
+Contrary to the visualization of the compositions within the ILR space as discussed in Section \ref{sec:balances}, here, one can analyses all parts of each compositions within a single plot.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/moreclasses/histo}
+  \caption{Shapley compositions visualized as histograms for the seven classes digit recognition example.}
+  \label{fig:histiris}
+\end{figure}
 
@@
 ...
+
 Features INDEPENDENCE!!
@@
 \label{app:algo}
-In this work, we used an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we present in this section.
+In this section, we present an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we used in this work for the estimation the Shapley compositions.
 
@@
 We want to optimally distribute the $m_{\text{max}}$ drawn samples over the $d$ features.
-Let $\hat{\bm{\phi}_i}$ the estimation of the Shapley composition for the $i$th feature.
-We want to minimize the sum of squared errors or distances $\displaystyle\sum_{i=1}^{d} \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a^2$.
+Let $\hat{\bm{\phi}_i}$ be the estimation of the Shapley composition for the $i$th feature.
+We want to minimize the sum of squared errors: $\displaystyle\sum_{i=1}^{d} \| \hat{\bm{\phi}_i} \ominus \bm{\phi}_i \|_a^2$.
 Since $\hat{\bm{\phi}_i}$ is a (Aitchison) sample mean we have: $\tilde{\hat{\bm{\phi}_i}} \approx \mathcal{N}(\tilde{\bm{\phi}}_i, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ and $\tilde{\hat{\bm{\phi}_i}} - \tilde{\bm{\phi}}_i \approx \mathcal{N}(\bm{0}, \frac{1}{m_i}\bm{\Sigma}^{(i)})$ where the tilde refers to the ILR transformation.

### 2023-12-15 shapleycomposition.py

@@ -213,11 +213,8 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
             for i in range(self.n_class):
                 fig.add_trace(go.Scatter3d(x=[0,class_vect[i,balances[0]-1]], y=[0,class_vect[i,balances[1]-1]],z=[0,class_vect[i,balances[2]-1]], mode='lines', line={'dash': 'dash', 'width' : 5}, name=self.names_classes[i], legendgroup='class', legendgrouptitle_text='Class composition:'))
 
-
         #PLOT THE SHAPLEY COMPOSITION IN THE ILR (SUB)SPACE
-        
         #If shapley_sum is True, the sum of the shapley vectors are summed fro the base distribution to the prediction
-
         if shapley_sum:
             if len(balances) == 2:
                 s = self.base.copy()
@@ -252,3 +249,16 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         fig.show()
 
         return fig
+
+    def shapley_histogram(self, figwidth=500, figheight=400):
+        fig = go.Figure(layout=go.Layout(autosize=False, width=figwidth, height=figheight))
+        fig.update_layout(barmode='group', legend=dict(orientation='h', bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="right", x=1, title='Classes'), margin=dict(l=0, r=0, t=0, b=0))
+        for i in range(self.n_class):
+            fig.add_trace(go.Bar(name=self.names_classes[i], x=self.names_features, y=ilr_inv(self.shapley, basis=self.basis)[:,i]))
+        fig.update_layout(bargroupgap=0)
+        fig.show()
+        
+        return fig
+
+
+    

### 2023-12-19 paper/main.tex
edit paper

@@
 When there are more classes, all the dimension of the ILR space cannot be visualized at once but $2$ or $3$-dimensional subspaces can still be visualized.
-However, in order to choose which ILR components to visualize, one needs to understand what they refer too.
-The next section provides some intuition about that in the context of the explanation of a classifier prediction with more than 4 classes.
+However, in order to choose which ILR components to visualize, one needs to understand what they refer to.
+For this purpose, the next section discusses the interpretation of the ILR components in more details in the context of the explanation of a classifier prediction with more than 4 classes.
 

### 2023-12-19 shapleycomposition.py

@@ -174,10 +174,9 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         if len(balances) == 2:
             fig.update_xaxes(range=[-lim, lim])
             fig.update_yaxes(range=[-lim, lim])
-            fig.update_layout(xaxis_title="ILR"+str(balances[0]), yaxis_title="ILR"+str(balances[1]), legend=dict(bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="right", x=1), font=dict(size=10), margin=dict(l=0, r=0, t=0, b=0))
+            fig.update_layout(xaxis_title="ILR"+str(balances[0]), yaxis_title="ILR"+str(balances[1]), legend=dict(bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="right", x=1), margin=dict(l=0, r=0, t=0, b=0))
         else:
             fig.update_layout(legend=dict(bgcolor='rgba(255,255,255,0.4)', yanchor="top", y=0.99, xanchor="left", x=0.01),
-                              font=dict(size=10),
                               margin=dict(l=0, r=0, t=0, b=0),
                               scene = dict(xaxis = dict(title="ILR"+str(balances[0]), range=[-lim,lim]),
                                            yaxis = dict(title="ILR"+str(balances[1]), range=[-lim,lim]),

### 2023-12-19 paper/bifurctree.tex
edit paper

\begin{tikzpicture}[scale=1]
  \draw (0,0) -- (1.1,1.1);
  \draw[dashed] (1.1,1.1) -- (1.4,1.4);
  \draw (1.4,1.4) -- (1.5,1.5);
  \draw (0.5,0.5) -- (1,0);
  \draw (1,1) -- (2,0);
  \draw (1.5,1.5) -- (3,0);

  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{1}$};
  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{2}$};
  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{N-1}$};

  \draw[black] (0,-0.25) node{$p_1$};
  \draw[black] (1,-0.25) node{$p_2$};
  \draw[black] (2,-0.25) node{$p_3$};
  \draw[black] (3,-0.25) node{$p_{N}$};
\end{tikzpicture}

### 2023-12-19 paper/main.tex
edit paper

--- 2023-12-15 paper/main.tex
@@
 \usepackage{microtype}
+\usepackage{tikz}
+\usepackage{pgfplots}
 \usepackage{graphicx}
@@
 An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built.
-The projection of a composition (a discrete probability distribution in our case) into this basis defined an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
+The projection of a composition (like a discrete probability distribution) into this basis defines an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
 This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry.
@@
 
-Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express is ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
+Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
 The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
-The choice of the basis will be discussed in Section \ref{sec:moreclasses}.
+The choice of the basis will be discussed in Section \ref{sec:balances}.
 
@@
 In a four classes example, the simplex is $3$-dimensional.
-We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}.
+We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}.
 It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3.
@@
 
-\subsubsection{More classes}
-\label{sec:moreclasses}
-When there are more classes, all the dimension of the ILR space cannot be visualized at once but $2$ or $3$-dimensional subspaces can still be visualized.
-However, in order to choose which ILR components to visualize, one needs to understand what they refer too.
-The next section provides some intuition about that in the context of the explanation of a classifier prediction with more than 4 classes.
+\subsection{More classes: groups of parts and balances}
+\label{sec:balances}
 
-\subsection{Groups of part, balances and sequential binary partition}
-\ref{sec:balances}
-..TODO..
+When there are more classes, all the dimension of the ILR space cannot be visualized at once.
+However, $2$ or $3$-dimensional subspaces can still be visualized.
+In order to choose which ILR components to visualize, one needs to understand what they refer to.
+In this section, we briefly discuss the interpretation of the ILR components in the context of the explanation of a classifier prediction with more than 4 classes.
+
+The projection of a composition into an element of the Aitchison orthonormal basis, i.e.
+a component of the ILR space.
+This can be interpreted as a \emph{balance}, i.e.
+a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}.
+One giving the central values of the parts in one group of parts and one for another group of parts.
+Therefore, in our context, a balances is comparing the weight of two group of classes.
+This can be illustrated by a sequential binary partition of bifurcation tree.
+Figure \ref{fig:trees} gives two examples.
+\begin{figure}
+  \centering
+  \subfigure[]{\label{fig:bifurc1}
+    \input{bifurctree.tex}
+    }
+  \subfigure[]{\label{fig:bifurc2}
+    \input{bifurctree.tex}
+  }
+  \caption{Two examples of bifurcation tree.
+\ref{fig:bifurc1} is the one corresponding to the basis obtained by with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}...}
+  \label{fig:trees}
+\end{figure}
+
+Geometric orthogonality (not statistical uncorrelation), balances provide nonredundant information about the composition.\\
+Balancing elements\\
+\cite{egozcue2005groups}
 
@@
 It gives the overall contribution of the feature on the prediction, regardless of its direction.
-The angle between two Shapley compositions can informs if features are redundant, orthogonal or contradictory.
+The angle between two Shapley compositions can informs about their orthogonality.
+If two Shapley composition are orthogonal\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}, this suggest the features are nonredundant.
+A negative angle would suggest that the features have an opposite influence on the prediction.
 The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing.

### 2023-12-20 paper/bifurctree1.tex
edit paper and code

\begin{tikzpicture}[scale=1]
  \draw (0,0) -- (1.5,1.5);
  \draw (0.25,0.25) -- (0.5,0);
  \draw (0.5,0.5) -- (1,0);
  \draw (0.75,0.75) -- (1.5,0);
  \draw (1,1) -- (2,0);
  \draw (1.5,1.5) -- (3,0);
  \draw (2.75,0.25) -- (2.5,0);

  \filldraw[black] (0.25,0.25) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{2}$};
  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{3}$};
  \filldraw[black] (0.75,0.75) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{4}$};
  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{5}$};
  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{6}$};
  \filldraw[black] (2.75,0.25) circle (0.75pt) node[anchor=south west]{\small $\tilde{p}_{1}$};

  \draw[black] (0,-0.25) node{$p_6$};
  \draw[black] (0.5,-0.25) node{$p_7$};
  \draw[black] (1,-0.25) node{$p_5$};
  \draw[black] (1.5,-0.25) node{$p_2$};
  \draw[black] (2,-0.25) node{$p_1$};
  \draw[black] (2.5,-0.25) node{$p_3$};
  \draw[black] (3,-0.25) node{$p_4$};
\end{tikzpicture}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2023-12-20 paper/main.tex
edit paper and code

@@
 
-The projection of a composition into an element of the Aitchison orthonormal basis, i.e.
-a component of the ILR space.
-This can be interpreted as a \emph{balance}, i.e.
-a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}.
-One giving the central values of the parts in one group of parts and one for another group of parts.
-Therefore, in our context, a balances is comparing the weight of two group of classes.
-This can be illustrated by a sequential binary partition of bifurcation tree.
+A component of the ILR space  can be interpreted as a \emph{balance}, i.e.
+a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes.
+Therefore, a balance is here comparing the weight of two group of classes.
+This can be illustrated by a sequential binary partition or bifurcation tree.
 Figure \ref{fig:trees} gives two examples.
+\ref{fig:bifurc1} shows the bifurcation tree corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
+The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$.
+Each next balance then recursively compares the probability for the next classe with the probabilities for the previous ones.
+\ref{fig:bifurc2} shows another bifurcation tree....
 \begin{figure}
@@
   \subfigure[]{\label{fig:bifurc2}
-    \input{bifurctree.tex}
+    \input{bifurctree1.tex}
   }
   \caption{Two examples of bifurcation tree.
-\ref{fig:bifurc1} is the one corresponding to the basis obtained by with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}...}
+\ref{fig:bifurc1} corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum} is another example corresponding to the basis used in the example...}
   \label{fig:trees}
@@
 Some may find the fine analysis of the features contributions in cases with more than four classes tricky.
-Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition.
+Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition.
 However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections.

### 2023-12-20 bifurc_tree.py

@@ -42,9 +42,11 @@ def create_tree_from_sbp(S, Np, depth=1, p=None):
     else:
         if S[1,colr][0] == 0:
             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[len(cols):,colr], Np, depth=depth+len(cols),p=r), create_tree_from_sbp(S[1:len(cols),cols], Np, depth=depth+1,p=s))
-        else:    
+        else:
             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[1:len(colr),colr], Np, depth=depth+1,p=r), create_tree_from_sbp(S[len(colr):,cols], Np, depth=depth+len(colr),p=s))
     return root
+
+
 
 def init_graph():
     return graphviz.Digraph(strict=True)

### 2023-12-20 shapleycomposition.py

@@ -176,7 +176,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
             fig.update_yaxes(range=[-lim, lim])
             fig.update_layout(xaxis_title="ILR"+str(balances[0]), yaxis_title="ILR"+str(balances[1]), legend=dict(bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="right", x=1), margin=dict(l=0, r=0, t=0, b=0))
         else:
-            fig.update_layout(legend=dict(bgcolor='rgba(255,255,255,0.4)', yanchor="top", y=0.99, xanchor="left", x=0.01),
+            fig.update_layout(legend=dict(bgcolor='rgba(255,255,255,0.4)', yanchor="top", y=0.99, xanchor="right", x=1),
                               margin=dict(l=0, r=0, t=0, b=0),
                               scene = dict(xaxis = dict(title="ILR"+str(balances[0]), range=[-lim,lim]),
                                            yaxis = dict(title="ILR"+str(balances[1]), range=[-lim,lim]),

### 2023-12-21 paper/main.tex
update tree guild

@@
 \twocolumn[
-\icmltitle{Explaining probabilistic prediction on the simplex with Shapley compositions}
+\icmltitle{Explaining probabilistic predictions on the simplex with Shapley compositions}
 
@@
 \begin{abstract}
-  The concept of Shapley value has been widely use for measuring the contribution of each feature on a machine learning model's prediction.
-However, this has been designed for one-dimensional function's codomain.
-For multiclass probabilistic classifier, where the output is a discrete probability distribution over the set of more than two possible classes, the output lives on a multidimensional simplex.
-In this case, people have been applying the concept of Shapley value on each output dimension one-by-one, in an implicit one-vs-rest setting, ignoring the compositional nature of the output distribution where the relative information between probabilities matter.
-Using the Aitchison geometry of the simplex, coming from the field of compositional data analysis, this paper present a first initiative for a multidimensional extention of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions on the simplex in machine learning.
+  The Shapley value has been widely used for measuring the contribution of each feature to a model's prediction.
+However, coming from game theory, this has been designed for a one-dimensional function's codomain.
+For a multiclass probabilistic classifier, the output is a discrete probability distribution, over a set of more than two possible classes, and lives on a multidimensional simplex.
+In this case, the Shapley values are sometimes computed on each output dimension one-by-one, in an implicit one-vs-rest setting, ignoring the compositional nature of the output distribution.
+Indeed, elements of the simplex are known as compositional data and a discrete probability distribution can therefore be treated as such taking into account the relative information between probabilities.
+Using the Aitchison geometry of the simplex, this paper presents a first initiative for a multidimensional extension of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions on the simplex in machine learning.
 \end{abstract}

### 2023-12-21 bifurc_tree.py

@@ -21,32 +21,71 @@ class Node:
         print(")",end="")
 
 
-def create_tree_from_sbp(S, Np, depth=1, p=None):
+# def create_tree_from_sbp(S, Np, depth=1, p=None):
+#     #check dim of the matrix
+#     if S.shape[0] != S.shape[1]-1: raise NameError('This is not a proper sequential binary partition matrix')
+
+#     if p is None:
+#         p = np.linspace(1,Np,Np, dtype=int)
+#         if S.shape[1] != Np: raise NameError('The sequential binary parition matrix does not match the number of parts in the composition')
+
+#     print('yolo')
+#     print(S)
+#     colr = np.argwhere(S[0,:]==1).squeeze(axis=1)
+#     cols = np.argwhere(S[0,:]==-1).squeeze(axis=1)
+#     print(colr)
+#     print(cols)
+#     r = p[colr]
+#     s = p[cols]
+#     if len(colr) == 1 and len(cols) == 1:
+#         print('1')
+#         root = Node("ilr", Np-depth, list(r)+list(s), Node("p", r[0], list(r)), Node("p", s[0], list(s)))
+#     elif len(colr) == 1:
+#         print('2')
+#         root = Node("ilr", Np-depth, list(r)+list(s), Node("p", r[0], list(r)), create_tree_from_sbp(S[1:,cols], Np, depth=depth+1, p=s))
+#     elif len(cols) == 1:
+#         print('3')
+#         root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[1:len(colr),colr], Np, depth=depth+1,p=r), Node("p", s[0], list(s)))
+#     else:
+#         if S[1,colr][0] == 0:
+#             print('4')
+#             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[len(cols):,colr], Np, depth=depth+len(cols),p=r), create_tree_from_sbp(S[1:len(cols),cols], Np, depth=depth+1,p=s))
+#         else:
+#             print('5')
+#             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[1:len(colr),colr], Np, depth=depth+1,p=r), create_tree_from_sbp(S[len(colr):,cols], Np, depth=depth+len(colr),p=s))
+#     return root
+
+def create_tree_from_sbp(S, Np):
     #check dim of the matrix
[…]

### 2023-12-21 mahalanobis_matrix.py

@@ -1,0 +1,19 @@
+import numpy as np
+from math import sqrt
+from composition_stats import ilr
+
+def mahalanobis_matrix(model, X, Y):
+    N = len(np.unique(Y))
+    P = ilr(model(X))
+    M = np.zeros((N, N))
+    for i in range(N):
+        for j in range(i+1,N):
+            Pi = P[Y == i]
+            Pj = P[Y == j]
+            mi = Pi.mean(axis=0)
+            mj = Pj.mean(axis=0)
+            co = ((len(Pi)-1)*np.cov(Pi, rowvar=False) + (len(Pj)-1)*np.cov(Pj, rowvar=False))/(len(Pi)+len(Pj)-2)
+            d  = sqrt(np.dot(np.dot(mi - mj , np.linalg.inv(co)),mi - mj))
+            M[i,j] = d
+            M[j,i] = d
+    return M

### 2023-12-21 more_classes_example.py

@@ -1,17 +1,22 @@
+# %%
+
 from sklearn.model_selection import train_test_split
 from sklearn import datasets, svm
 from sklearn.decomposition import PCA
 from sklearn.preprocessing import scale
+from sklearn.metrics import confusion_matrix
+from sklearn.cluster import AgglomerativeClustering
 import numpy as np
 from math import sqrt
 from shapleycomposition import ShapleyExplainer
 from composition_stats import ilr, sbp_basis
 from bifurc_tree import create_tree_from_sbp, init_graph, build_graph
 import plotly.graph_objects as go
+from mahalanobis_matrix import mahalanobis_matrix
 
 # %%
 K = 10         #Index of the instance you want to test in the test set
-N_class = 6    #Number of class, the dimension of the simplex is therefore N_class-1
+N_class = 10    #Number of class, the dimension of the simplex is therefore N_class-1
 N_feat  = 6    #In this example, since the number of feature of the digit dataset is quite large (64), we propose to reduce it with a PCA
 
 #load the dataset, take a subset of N_class classes, scale it and split into a training and testing set
@@ -34,10 +39,27 @@ svc_linear.fit(X_train, Y_train)
 
 # %%
 
+M = mahalanobis_matrix(svc_linear.predict_proba, X_test, Y_test)
+hie_clust = AgglomerativeClustering(affinity='precomputed', linkage='average').fit(M)
+child = hie_clust.children_
+
+
+# %%
+
 #Choose a sequential binary partition matrix and plot the corresponding bifurcation tree
-sbpmatrix = np.array([[1,1,-1,-1,1,1],[1,-1,0,0,-1,-1],[0,1,0,0,-1,-1],[0,0,0,0,1,-1],[0,0,1,-1,0,0]])
+sbpmatrix = np.array([[1,1,-1,-1,1,1,1],[1,-1,0,0,-1,-1,-1],[0,1,0,0,-1,-1,-1],[0,0,0,0,1,-1,-1],[0,0,0,0,0,1,-1],[0,0,1,-1,0,0,0]])
+#sbpmatrix = np.array([[1,-1,1,-1,1,-1,1],[1,0,1,0,-1,0,-1],[0,0,0,0,1,0,-1],[1,0,-1,0,1,0,0],[0,1,0,1,0,-1,0],[0,1,0,-1,0,0,0]])
+sbpmatrix = np.array([[0,0,0,0,0,0,0,1,-1,0],
[…]

### 2023-12-22 bifurc_tree.py
child to sbpm

@@ -20,41 +20,7 @@ class Node:
         else: raise NameError('This is not a proper bifurcation tree')
         print(")",end="")
 
-
-# def create_tree_from_sbp(S, Np, depth=1, p=None):
-#     #check dim of the matrix
-#     if S.shape[0] != S.shape[1]-1: raise NameError('This is not a proper sequential binary partition matrix')
-
-#     if p is None:
-#         p = np.linspace(1,Np,Np, dtype=int)
-#         if S.shape[1] != Np: raise NameError('The sequential binary parition matrix does not match the number of parts in the composition')
-
-#     print('yolo')
-#     print(S)
-#     colr = np.argwhere(S[0,:]==1).squeeze(axis=1)
-#     cols = np.argwhere(S[0,:]==-1).squeeze(axis=1)
-#     print(colr)
-#     print(cols)
-#     r = p[colr]
-#     s = p[cols]
-#     if len(colr) == 1 and len(cols) == 1:
-#         print('1')
-#         root = Node("ilr", Np-depth, list(r)+list(s), Node("p", r[0], list(r)), Node("p", s[0], list(s)))
-#     elif len(colr) == 1:
-#         print('2')
-#         root = Node("ilr", Np-depth, list(r)+list(s), Node("p", r[0], list(r)), create_tree_from_sbp(S[1:,cols], Np, depth=depth+1, p=s))
-#     elif len(cols) == 1:
-#         print('3')
-#         root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[1:len(colr),colr], Np, depth=depth+1,p=r), Node("p", s[0], list(s)))
-#     else:
-#         if S[1,colr][0] == 0:
-#             print('4')
-#             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[len(cols):,colr], Np, depth=depth+len(cols),p=r), create_tree_from_sbp(S[1:len(cols),cols], Np, depth=depth+1,p=s))
-#         else:
-#             print('5')
-#             root = Node("ilr", Np-depth, list(r)+list(s), create_tree_from_sbp(S[1:len(colr),colr], Np, depth=depth+1,p=r), create_tree_from_sbp(S[len(colr):,cols], Np, depth=depth+len(colr),p=s))
-#     return root
-
+        
 def create_tree_from_sbp(S, Np):
     #check dim of the matrix
[…]

### 2023-12-22 more_classes_example.py

@@ -10,13 +10,13 @@ import numpy as np
 from math import sqrt
 from shapleycomposition import ShapleyExplainer
 from composition_stats import ilr, sbp_basis
-from bifurc_tree import create_tree_from_sbp, init_graph, build_graph
+from bifurc_tree import create_tree_from_sbp, init_graph, build_graph, sbp_from_aggloclustchildren
 import plotly.graph_objects as go
 from mahalanobis_matrix import mahalanobis_matrix
 
 # %%
 K = 10         #Index of the instance you want to test in the test set
-N_class = 10    #Number of class, the dimension of the simplex is therefore N_class-1
+N_class = 10   #Number of class, the dimension of the simplex is therefore N_class-1
 N_feat  = 6    #In this example, since the number of feature of the digit dataset is quite large (64), we propose to reduce it with a PCA
 
 #load the dataset, take a subset of N_class classes, scale it and split into a training and testing set
@@ -39,25 +39,14 @@ svc_linear.fit(X_train, Y_train)
 
 # %%
 
-M = mahalanobis_matrix(svc_linear.predict_proba, X_test, Y_test)
-hie_clust = AgglomerativeClustering(affinity='precomputed', linkage='average').fit(M)
-child = hie_clust.children_
-
-
-# %%
-
 #Choose a sequential binary partition matrix and plot the corresponding bifurcation tree
-sbpmatrix = np.array([[1,1,-1,-1,1,1,1],[1,-1,0,0,-1,-1,-1],[0,1,0,0,-1,-1,-1],[0,0,0,0,1,-1,-1],[0,0,0,0,0,1,-1],[0,0,1,-1,0,0,0]])
+# sbpmatrix = np.array([[1,1,-1,-1,1,1,1],[1,-1,0,0,-1,-1,-1],[0,1,0,0,-1,-1,-1],[0,0,0,0,1,-1,-1],[0,0,0,0,0,1,-1],[0,0,1,-1,0,0,0]])
 #sbpmatrix = np.array([[1,-1,1,-1,1,-1,1],[1,0,1,0,-1,0,-1],[0,0,0,0,1,0,-1],[1,0,-1,0,1,0,0],[0,1,0,1,0,-1,0],[0,1,0,-1,0,0,0]])
-sbpmatrix = np.array([[0,0,0,0,0,0,0,1,-1,0],
-                      [0,0,0,1,0,0,0,0,0,-1],
[…]

### 2023-12-24 paper/bifurctree1.tex
edit

@@
-\begin{tikzpicture}[scale=1]
-  \draw (0,0) -- (1.5,1.5);
-  \draw (0.25,0.25) -- (0.5,0);
-  \draw (0.5,0.5) -- (1,0);
-  \draw (0.75,0.75) -- (1.5,0);
-  \draw (1,1) -- (2,0);
-  \draw (1.5,1.5) -- (3,0);
-  \draw (2.75,0.25) -- (2.5,0);
+\begin{tikzpicture}[scale=1.7]
+  \draw (0,0) -- (2.25,2.25);
+  \draw (2.25,2.25) -- (4.5,0);
+  \draw (2,1.5) -- (0.5,0);
+  \draw (1.25,0.25) -- (1,0);
+  \draw (1,0.5) -- (1.5,0);
+  \draw (3.25,0.25) -- (3,0);
+  \draw (4.25,0.25) -- (4,0);
+  \draw (1.75,1.75) -- (3.5,0);
+  \draw (3,0.5) -- (2.5,0);
+  \draw (2.75,0.75) -- (2,0);
 
-  \filldraw[black] (0.25,0.25) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{2}$};
-  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{3}$};
-  \filldraw[black] (0.75,0.75) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{4}$};
-  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{5}$};
-  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\small $\tilde{p}_{6}$};
-  \filldraw[black] (2.75,0.25) circle (0.75pt) node[anchor=south west]{\small $\tilde{p}_{1}$};
+  \filldraw[black] (1.75,1.75) circle (0.75pt) node[label={[anchor=south east]0.1mm:$\tilde{p}_{8}$}]{};
+  \filldraw[black] (3.25,0.25) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{2}$};
+  \filldraw[black] (4.25,0.25) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{3}$};
+  \filldraw[black] (1.25,0.25) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{1}$};
+  \filldraw[black] (1,0.5) circle (0.75pt) node[label={[anchor=south east]0.1mm:\small $\tilde{p}_{5}$}]{};
 
-  \draw[black] (0,-0.25) node{$p_6$};
-  \draw[black] (0.5,-0.25) node{$p_7$};
-  \draw[black] (1,-0.25) node{$p_5$};
-  \draw[black] (1.5,-0.25) node{$p_2$};
-  \draw[black] (2,-0.25) node{$p_1$};
-  \draw[black] (2.5,-0.25) node{$p_3$};
-  \draw[black] (3,-0.25) node{$p_4$};
+  \filldraw[black] (2,1.5) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{7}$};
+  \filldraw[black] (3,0.5) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{4}$};
+  \filldraw[black] (2.75,0.75) circle (0.75pt) node[anchor=south west]{$\tilde{p}_{6}$};
+
+  \draw[black] (0,-0.25) node{$4$};
+  \draw[black] (0.5,-0.25) node{$1$};
+  \draw[black] (1,-0.25) node{$7$};
+  \draw[black] (1.5,-0.25) node{$8$};
+  \draw[black] (2,-0.25) node{$5$};
+  \draw[black] (2.5,-0.25) node{$2$};
+  \draw[black] (3,-0.25) node{$3$};
+  \draw[black] (3.5,-0.25) node{$9$};
+  \draw[black] (4,-0.25) node{$0$};
+  \draw[black] (4.5,-0.25) node{$6$};
 \end{tikzpicture}

### 2023-12-24 paper/main.tex
edit

@@
 \usepackage{graphicx}
-\usepackage{subfigure}
+\usepackage{subcaption}
 \usepackage{booktabs} % for professional tables
@@
 \begin{figure}
+
+  \begin{subfigure}{0.5\textwidth}
     \centering
-  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
     \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
-    }
-  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
+    \caption{Shapley compositions in the ILR space.}
+    \label{fig:3classesshap}
+  \end{subfigure}
+
+\vspace{0.5cm}
+  \begin{subfigure}{0.5\textwidth}
+    \centering
     \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
-  }
+    \caption{Sum of the Shapley compositions in the ILR space from the base to the prediction.}
+    \label{fig:3classesshapsum}
+  \end{subfigure}
+
   \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
@@
 \begin{figure}
+  \begin{subfigure}{0.5\textwidth}
     \centering
-  \subfigure[]{\label{fig:bifurc1}
     \input{bifurctree.tex}
-    }
-  \subfigure[]{\label{fig:bifurc2}
+    \caption{Bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} and used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}.}
+    \label{fig:bifurc1}
+  \end{subfigure}
+
+  \vspace{0.5cm}
+  \begin{subfigure}{0.5\textwidth}
+      \centering
       \input{bifurctree1.tex}
-  }
-  \caption{Two examples of bifurcation tree.
-\ref{fig:bifurc1} corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum} is another example corresponding to the basis used in the example...}
+      \caption{Bifurcation tree used for our $10$-class recognition task}
+      \label{fig:bifurc2}
+    \end{subfigure}
+  \caption{Two examples of bifurcation tree. \ref{fig:bifurc1}.}
   \label{fig:trees}

### 2023-12-24 more_classes_example.py

@@ -45,7 +45,7 @@ svc_linear.fit(X_train, Y_train)
 
 #Get a sequential binary parition from the agglomeration of classes from a distance matrix between classes. Distance matrix is here made of the mahalanobis distance between a pair of classes in the ILR space output by the classifier (assuming normal distribution with same covariance matrix between a pair of classes).
 M = mahalanobis_matrix(svc_linear.predict_proba, X_test, Y_test)
-sbpmatrix = sbp_from_aggloclustchildren(AgglomerativeClustering(affinity='precomputed', linkage='average').fit(M).children_)
+sbpmatrix = sbp_from_aggloclustchildren(AgglomerativeClustering(metric='precomputed', linkage='average').fit(M).children_)
 
 basis = np.flip(sbp_basis(sbpmatrix), axis=0)
 root = create_tree_from_sbp(sbpmatrix, N_class)

### 2023-12-30 paper/main.tex
edit paper and code

@@
 
-When there are more classes, all the dimension of the ILR space cannot be visualized at once.
-However, $2$ or $3$-dimensional subspaces can still be visualized.
-In order to choose which ILR components to visualize, one needs to understand what they refer to.
-In this section, we briefly discuss the interpretation of the ILR components in the context of the explanation of a classifier prediction with more than 4 classes.
+When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once, but $2$ or $3$-dimensional subspaces can still be visualized.
+In order to select the ILR components to visualize, one needs to understand what they refer to.
+In this section, we briefly discuss the interpretation of the ILR components in our context.% of the explanation of a classifier prediction with more than three classes.
 
 A component of the ILR space  can be interpreted as a \emph{balance}, i.e.
-a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes.
-Therefore, a balance is here comparing the weight of two group of classes.
+a log-ratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes.
+Therefore, a balance is here comparing the weight of two groups of classes.
+The set of balances is built such that they are geometrically orthogonal meaning they provide nonredundant information\footnote{Not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}.
+
 This can be illustrated by a sequential binary partition or bifurcation tree.
 Figure \ref{fig:trees} gives two examples.
-\ref{fig:bifurc1} shows the bifurcation tree corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
+\ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
 The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$.
-Each next balance then recursively compares the probability for the next classe with the probabilities for the previous ones.
-\ref{fig:bifurc2} shows another bifurcation tree....
+Each next balance then recursively compares the probability for the next class with the probabilities for the previous ones independently of all the others.
+
+In some applications, one may be interested in particular comparisons of groups of classes.
+For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups.
+But sometimes, there are no obvious relevant comparisons to study.
+For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime.
+However, being basically a shape recognition problem and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent.
+Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy.
+Let's use the basis of Figure \ref{fig:bifurc2} for explaining the contribution of the features in a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
+Let say for conciseness that we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8.
+We therefore only look at the the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$) as in Figure \ref{fig:moreclasses35}.
+
 \begin{figure}
@@
       \input{bifurctree1.tex}
-      \caption{Bifurcation tree used for our $10$-class recognition task}
+      \caption{Bifurcation tree used for our $10$-classes digit recognition task}
       \label{fig:bifurc2}
@@
 \end{figure}
-
-Geometric orthogonality (not statistical uncorrelation), balances provide nonredundant information about the composition.\\
-Balancing elements\\
-\cite{egozcue2005groups}
 
@@
     \cline{2-9}
-    & \multirow{2}{*}{norm} & \multicolumn{3}{c|}{projection on the class-compositions} & \multicolumn{4}{c|}{cosine similarity between each Shapley composition} \\
+    & \multirow{2}{*}{norm} & \multicolumn{3}{c|}{projection on the class-compositions} & \multicolumn{4}{c|}{cosine similarity between Shapley compositions} \\
     \cline{3-9}
-     & & setosa & versicolor & virginica & pet. length & pet. width & sep. length & sep. width \\
+     & & \small setosa & \small versicolor & \small virginica & \small pet. length & \small pet. width & \small sep. length & \small sep. width \\
     \hline
-    pet. length & 2.20  & -1.91 & 1.90 & 0.02 & 1 &$\cdot$ &$\cdot$ &$\cdot$ \\
-    pet. width & 0.51 & -0.51 & 0.29 & 0.22 & 0.90 & 1 &$\cdot$ &$\cdot$ \\
-    sep. length & 0.13 & -0.08 & 0.13 & -0.05 & 0.93 & 0.69 & 1 &$\cdot$ \\
-    sep. width & 0.11 & -0.11 & 0.04 & 0.07 & 0.77 & 0.97 & 0.49 &1 \\
+    \small pet. length & 2.20  & -1.91 & 1.90 & 0.02 & 1 &$\cdot$ &$\cdot$ &$\cdot$ \\
+    \small pet. width & 0.51 & -0.51 & 0.29 & 0.22 & 0.90 & 1 &$\cdot$ &$\cdot$ \\
+    \small sep. length & 0.13 & -0.08 & 0.13 & -0.05 & 0.93 & 0.69 & 1 &$\cdot$ \\
+    \small sep. width & 0.11 & -0.11 & 0.04 & 0.07 & 0.77 & 0.97 & 0.49 &1 \\
     \hline
@@
 In order to respect the efficiency property, we want this error to be the neutral element of the perturnation, i.e.
-the ``zero'' in the sense of the Aitchison geometry, i.e the uniform distribution.
+the ``zero'' in the sense of the Aitchison geometry: the uniform distribution.
 We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.
@@
 \end{equation}
-In this way, when $\epsilon$ goes to zero, the efficiency property is respected for the adjusted Shapley compositions and more weight is given to the adjusments of the Shapley compositions with a higher estimation variance.
+In this way, when $\epsilon$ goes to zero\footnote{In our experiments, $\epsilon = 10^{-6}$.}, the efficiency property is respected for the adjusted Shapley compositions and more weight is given to the adjusments of the Shapley compositions with a higher estimation variance.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

### 2023-12-30 bifurc_tree.py

@@ -97,6 +97,7 @@ def sbp_from_aggloclustchildren(children):
         else:
             S[i,M[children[i,1]-children.shape[0]-1]] = -1
             M[i] = M[i]+M[children[i,1]-children.shape[0]-1]
+    S = S[np.argsort((S!=0).sum(axis=1)),:]
     return S
         
 # S = np.array([[1,1,-1,-1,1,1],[1,-1,0,0,-1,-1],[0,1,0,0,-1,-1],[0,0,0,0,1,-1],[0,0,1,-1,0,0]])

### 2023-12-30 shapleycomposition.py

@@ -38,8 +38,15 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         self.class_compo = class_compositions(n_class)   #class_compo[k,:] is the kth class composition meaning a distribution going straigth in favor of class k with an unit norm.
         if sbpmatrix is None:
             self.basis = None
+            self.sbpmatrix = np.zeros((n_class-1, n_class))
+            for i in range(n_class-1):
+                self.sbpmatrix[i,[j for j in range(i+1)]] = 1
+                self.sbpmatrix[i,i+1] = -1
+            print(self.sbpmatrix)
         else:
-            self.basis = np.flip(sbp_basis(sbpmatrix), axis=0)
+            #self.basis = np.flip(sbp_basis(sbpmatrix), axis=0)
+            self.basis = sbp_basis(sbpmatrix)
+            self.sbpmatrix = sbpmatrix
         self.base      = ilr(model(train_data), basis=self.basis).mean(axis=0)
         self.pred      = None
         self.shapley   = None
@@ -174,7 +181,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
         if len(balances) == 2:
             fig.update_xaxes(range=[-lim, lim])
             fig.update_yaxes(range=[-lim, lim])
-            fig.update_layout(xaxis_title="ILR"+str(balances[0]), yaxis_title="ILR"+str(balances[1]), legend=dict(bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="right", x=1), margin=dict(l=0, r=0, t=0, b=0))
+            fig.update_layout(xaxis_title="ILR"+str(balances[0]), yaxis_title="ILR"+str(balances[1]), legend=dict(bgcolor='rgba(255,255,255,0.4)',yanchor="top", y=0.99, xanchor="left", x=0.01), margin=dict(l=0, r=0, t=0, b=0))
         else:
             fig.update_layout(legend=dict(bgcolor='rgba(255,255,255,0.4)', yanchor="top", y=0.99, xanchor="right", x=1),
                               margin=dict(l=0, r=0, t=0, b=0),
@@ -184,7 +191,7 @@ with feature contributions" by Erik Štrumbelj and Igor Kononenko.
 
 
[…]

### 2023-12-30 paper/main.tex
add tables

@@
 Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy.
-Let's use the basis of Figure \ref{fig:bifurc2} for explaining the contribution of the features in a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
-Let say for conciseness that we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8.
-We therefore only look at the the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$) as in Figure \ref{fig:moreclasses35}.
-
+Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
+We comment, for conciseness, a single $2$-dimensional subspace.
+More visualisation can be found in Appendix \ref{app:...}.
+Let's have a look at the the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$) as in Figure \ref{fig:moreclasses35}.
+It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8.
+$\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8.
+Therefore, the class-composition for the others digits have a zero projection within this subspace and are not drawn in Figure \ref{fig:moreclasses35}.
+The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8.
+Indeed, the groups of digits making the balance $\tilde{p}_3$ and the groups of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor.
+In this case, none of the class-compositions are orthogonal.
+In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection of the class-compositions on this line for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8.
+The two latters, are equal and are half as long as the former.
+In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.
 \begin{figure}
@@
       \input{bifurctree1.tex}
-      \caption{Bifurcation tree used for our $10$-classes digit recognition task}
+      \caption{Bifurcation tree used in our $10$-classes digit recognition task}
       \label{fig:bifurc2}
     \end{subfigure}
-  \caption{Two examples of bifurcation tree. \ref{fig:bifurc1}.}
+  \caption{Two examples of bifurcation tree.}
   \label{fig:trees}
 \end{figure}
-
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
+  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$.}
+  \label{fig:moreclasses35}
+\end{figure}
 \subsection{Angles, norms and projections}
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/3classes/histo}
+  \includegraphics[width=0.9\linewidth]{figures/3classes/histo}
   \caption{Shapley compositions visualized as histograms for the Iris classification example.}
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/moreclasses/histo}
+  \includegraphics[width=0.9\linewidth]{figures/moreclasses/histo}
   \caption{Shapley compositions visualized as histograms for the seven classes digit recognition example.}
@@
 
-The Table \ref{tab:normiris} gives, for the Iris classification example of Figure \ref{fig:3classes}, the norm of each Shapley compositions, the cosine similarity between them and their projection on the set of class-compositions.
-Having the highest norm, this confirms that the petal length is the feature that contributed the most on the prediction.
-Having a close to zero inner product with the virginica class-composition, shows that this feature did not contribute to the value of the probability of of this class.
-It contributes neither in favor, nor in the reject of this class.
-Having a positive inner product with the versicolor class-composition and a negative one with the setosa class-composition, this suggest that this feature is going in favor of the class versicolor and against setosa.
 \begin{table}
@@
 \end{table}
+The Table \ref{tab:normiris} gives, for the Iris classification example of Figure \ref{fig:3classes}, the norm of each Shapley compositions, the cosine similarity between them and their projection on the set of class-compositions.
+Having the highest norm, this confirms that the petal length is the feature that contributed the most on the prediction.
+Having a close to zero inner product with the virginica class-composition, shows that this feature did not contribute to the value of the probability of of this class.
+It contributes neither in favor, nor in the reject of this class.
+Having a positive inner product with the versicolor class-composition and a negative one with the setosa class-composition, this suggest that this feature is going in favor of the class versicolor and against setosa.
 
+\begin{table}
+  \centering
+  \caption{Norm of the Shapley compositions and Cosine similarity betweeen the Shapley compositions for the $10$-classes digit recognition example.}
+  \begin{tabular}{c|c|cccccc|}
+    \cline{2-8}
+    & \multirow{2}{*}{norm} & \multicolumn{6}{c|}{cosine similarity between Shapley compositions} \\
+    \cline{3-8}
+     & & \small feature n.1 & \small feature n.2 & \small feature n.3 & \small feature n.4 & \small feature n.5 & \small feature n.6 \\
+    \hline
+    \small feature n.1 & 4.00 & 1 & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ \\
+    \small feature n.2 & 1.15 & -0.17 & 1 & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ \\
+    \small feature n.3 & 2.25 & 0.24 & 0.23 & 1 & $\cdot$ & $\cdot$ &$\cdot$ \\
+    \small feature n.4 & 1.26 & -0.23 & 0.23 & 0.09 & 1 & $\cdot$ & $\cdot$ \\
+    \small feature n.5 & 0.60 & 0.18  & -0.17 & -0.01 & -0.56 & 1 & $\cdot$ \\
+    \small feature n.6 & 1.11 & 0.14 & 0.35 &	-0.19 & -0.03 & -0.48 & 1 \\
+    \hline
+  \end{tabular}
+  \label{tab:normiris}
+\end{table}
+
+\begin{table}
+  \centering
+  \caption{Projection of the Shapley compositions on the class-compositions for the $10$-classes digit recognition example.}
+  \begin{tabular}{c|cccccccccc|}
+    \cline{2-11}
+    & \multicolumn{10}{c|}{projection on the class-compositions} \\
+    \cline{2-11}
+    & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
+    \hline
+    \small feature n.1 & 1.65 & -0.85 & 0.09 & -0.57 & -1.92 & -0.34 & 1.74 & 1.96 & 0.22 & -1.81 \\
+    \small feature n.2 & -0.50 & 0.35 & 0.11 & 0.08 & -0.57 & -0.29 & -0.10 & 0.33 & -0.14 & 0.74\\
+    \small feature n.3 & 0.36 & 0.45 & 0.23 & 1.20 & -1.37 & -0.56 & 0.91 & -0.83 & -0.31 & -0.09 \\
+    \small feature n.4 & 0.19 & 0.02 & 0.61 & 0.18 & -0.53 & 0.46 & -0.86 & -0.17 & -0.12 & 0.22 \\
+    \small feature n.5 & -0.03 & -0.02 & -0.22 & -0.31 & -0.02 & 0.13 & 0.42 & -0.16& 0.14 & 0.07 \\
+    \small feature n.6 & -0.17 & 0.34 & -0.22 & 0.17 & -0.12 & -0.43 &-0.50  & 0.75 & 0.34 & -0.18 \\
+    \hline
+  \end{tabular}
+  \label{tab:normiris}
+\end{table}
 Other examples ... !!!

### 2024-01-03 paper/main.tex
edit paper

@@
 
-Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}$.
-Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{In practice, this is usually unknown but the expectation will be replaced by empirical samplings.}.
+Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}\subset\mathbb{R}^d$.
+Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{Usually, this is unknown but expectations will be replaced by empirical averagings.}.
 Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
@@
 When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$.
-However, when only $\bm{x}_S$ is given, i.e.
-part of the features, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
+However, when only $\bm{x}_S$ is given with $\mathcal{S} \neq \mathcal{I}$, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
 The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
@@
 \end{equation}
-where $2^{\mathcal{X}}$ is the set of all subsets of $I$.
+where $2^{\mathcal{I}}$ is the set of all subsets of $\mathcal{I}$.
 This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
@@
 \end{equation}
-where $\pi$ is a permutation of the set $I$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
+where $\pi$ is a permutation of the set $\mathcal{I}$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
 For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.
@@
 This quantity is known as the Shapley value for the $i$th feature.
-It comes from cooperative game theory and is known to be only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}.
-It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values: $(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i)$, which is known as the \emph{efficiency} property.
+It comes from cooperative game theory and is known to be the only quantity respecting a set of desired axiomatic properties \cite{shapley1953value}.
+It is linear as a function of the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$, and the ``centered'' learned model is additively separable with respect to the Shapley values:
+\begin{equation}
+  f(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i),
+\end{equation}
+which is known as the \emph{efficiency} property.
 
-Like originally developed in game theory, the Shapley value is designed for one-dimensinal codomain of the function $f$.
+The Shapley value is designed for one-dimensinal codomain of the function $f$.
 For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison.
-However, this ignores the relative information between each probability and the compositional nature of probability distributions.
+However, this approach ignores the relative information between each probability and ignores the compositional nature of probability distributions.
 Indeed, the probabilistic output of a classifier lives on a multidimensional simplex.
@@
 Each element of the vector is a part of the \emph{whole} $k$.
-The sample space of compositional data is known as the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$.
+The sample space of compositional data is the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$.
 In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}.
@@
 \item an inner product:\\
-  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$.
+  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$,
 \end{itemize}
-$\mathcal{C}(\cdot)$ is the closure operator.
+where $\mathcal{C}(\cdot)$ is the closure operator.
 Since only the relative information matter, scaling factors are irrelevant and a composition $\bm{x}$ is equivalent to $\lambda \bm{x} = [\lambda x_1,\lambda x_2,\dots\lambda x_N]$ for all $\lambda>0$.
@@
 Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
-The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
+The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where the set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
 The choice of the basis will be discussed in Section \ref{sec:balances}.
@@
 
-In Section \ref{sec:shapley} we have briefly presented the standard formulation of Shapley value designed for one-dimensional prediction.
-In this Section, we will see how the Aitchison geometry can be used to extend this concept to the multidimensional simplex for explaning probabilistic predictions.
+In Section \ref{sec:shapley} we have briefly presented the standard use of Shapley value designed for explaining one-dimensional predictions.
+In this Section, we will see how the Aitchison geometry can be used to extend the concept of Shapley value to the simplex for explaning multidimensional probabilistic predictions.
 
 Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$.
-To properly consider the relative information between the probabilities, The outputs of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex.
+To properly consider the relative information between the probabilities, the output of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex.
 We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
@@
 \end{equation}
-The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as \cite{pawlowskymodeling}: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
-where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure.
+The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
+where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure \cite{pawlowskymodeling}.
 
@@
 The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}.
-This space has the advantage of being intuitive since it is the standard real $(N-1)$-dimensional vector space we are used too.
+This space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.
 
@@
 Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction.
-Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region.
+Having the highest norm, the petal length is the feature contributing the most on the prediction, and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines).
 Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class.
@@
     \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
-    \caption{Sum of the Shapley compositions in the ILR space from the base to the prediction.}
+    \caption{Sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
     \label{fig:3classesshapsum}
@@
 It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3.
-Since they are 64 pixels as a set of features, which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for better clarity.
-We again use a SVM classifier with an rbf kernel.
+Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley composition.
+We therefore reduce the number of features to 6 using a principal component analysis for better clarity and conciseness.
+We again use a SVM classifier with a rbf kernel.
 The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}.
@@
 a $3$-dimensional simplex.
-The relative position of the class-compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
+The relative position of the class-compositions in the ILR space are the same as the bonds between the carbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
+In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double blind peer reviewing.}.
 \begin{figure}
@@
   \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task.
-The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
+The Shapley compositions are summed in the ILR space from the base distribution to the prediction.
+The gray transparent walls markout the maximum probability decision regions.}
   \label{fig:4classesshapsum}
@@
 In order to select the ILR components to visualize, one needs to understand what they refer to.
-In this section, we briefly discuss the interpretation of the ILR components in our context.% of the explanation of a classifier prediction with more than three classes.
+In this section, we briefly discuss the interpretation of the ILR components.% of the explanation of a classifier prediction with more than three classes.
 
@@
 This can be illustrated by a sequential binary partition or bifurcation tree.
-Figure \ref{fig:trees} gives two examples.
-\ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
+Figure \ref{fig:trees} gives two examples: \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
 The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$.
@@
 For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups.
-But sometimes, there are no obvious relevant comparisons to study.
-For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime.
-However, being basically a shape recognition problem and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent.
+However, there are sometimes no obvious relevant comparisons to study.
+For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent.
 Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy.
-Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
-We comment, for conciseness, a single $2$-dimensional subspace.
-More visualisation can be found in Appendix \ref{app:...}.
-Let's have a look at the the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$) as in Figure \ref{fig:moreclasses35}.
+Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
+We comment, for conciseness, a single $2$-dimensional subspace\footnote{More visualisations can be obtaned from the shared python notebooks.}.
+Let's have a look, in Figure \ref{fig:moreclasses35}, at the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$).
 It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8.
@@
 The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8.
-Indeed, the groups of digits making the balance $\tilde{p}_3$ and the groups of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor.
+Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor.
 In this case, none of the class-compositions are orthogonal.
@@
 In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.
+
+Within this space, Shapley compositions can be explored like in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum} keeping in mind that this is a subset of the full ILR space. 
 \begin{figure}
@@
   \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
-  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$.}
+  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from the class 2.
+$\tilde{p}_3$ compares the probability for the class 0 with the one for the class 6 and $\tilde{p}_5$ compares the probability for the class 1 with the group of probabilities for the class 7 and 8.
+}
   \label{fig:moreclasses35}
@@
 
-Some may find the fine analysis of the features contributions in cases with more than four classes tricky.
-Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition.
-However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections.
-Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction.
+Some may find the analysis of the features contributions in cases with more than four classes tricky.
+Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the balances and a careful building of the bifurcation tree.
+However, the Shapley explaination can be summarized by sets of angles, norms and projections.
+The norm of a Shapley composition gives the strength of the feature's contribution in the prediction.
 It gives the overall contribution of the feature on the prediction, regardless of its direction.
 The angle between two Shapley compositions can informs about their orthogonality.
-If two Shapley composition are orthogonal\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}, this suggest the features are nonredundant.
+If two Shapley composition are orthogonal%\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}
+, this suggest the features are nonredundant.
 A negative angle would suggest that the features have an opposite influence on the prediction.
@@
 In opposite, the histogram for the petal length as a high value for the versicolor class, relatively to the others, confirming the contribution of the feature toward this class.
+For the petal width, the value for the class setosa is low relatively to the others which confirms the constribution of this feature against the class setosa.
 \begin{figure}
@@
 \end{figure}
-As another illustration, Figure \ref{fig:histmore} shows the Shapley compositions of the seven classes digit recognition example.
+
+As another illustration, Figure \ref{fig:histmore} shows the Shapley compositions of the $10$-classes digit recognition example.
 Contrary to the visualization of the compositions within the ILR space as discussed in Section \ref{sec:balances}, here, one can analyses all parts of each compositions within a single plot.
+In this example, the high of the class 2 for the first principal component confirms the contribution of this features toward this class.
 \begin{figure}
@@
   \caption{Shapley compositions visualized as histograms for the seven classes digit recognition example.}
-  \label{fig:histiris}
+  \label{fig:histmore}
 \end{figure}
@@
 In this work, the estimation algorithm we used to compute the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}.
-Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement.
-Each Shapley compositions are therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold.
+Each Shapley composition is therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}} to respect the efficiency propery.
 See Appendix \ref{app:algo} and \ref{app:correct} for more details.
+
+\section{Shapley values versus Shapley composition}
+
+We just proposed a theoretically well founded extension of the concept of Shapley value for the explaining multidimensional prediction in machine learning.
+The use of standard Shapley value in this context has been rarely discuss in the literature.
+However, some compute a Shapley value on each output dimension one-by-one.
+We here briefly discuss and compare this approach with the Shapley composition framework.
+
+A Shapley value can be computed on the logit of the probability for each classes resuilting in $N$-dimensional vector of the Shapley values for a $N$-classes problem.
+Such vector will be refered here as \emph{Shapley vector}.
+Even if the efficiency property holds, i.e.
+the sum of the part-wise logit of the base distribution with the Shapley vectors for each feature is equal to the part-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
 
@@
   \centering
-  \caption{Norm of the Shapley compositions and Cosine similarity betweeen the Shapley compositions for the $10$-classes digit recognition example.}
+  \caption{Norm of the Shapley compositions and Cosine similarity betweeen the Shapley composition for the $10$-classes digit recognition example.}
   \begin{tabular}{c|c|cccccc|}
@@
   \end{tabular}
-  \label{tab:normiris}
+  \label{tab:normdigit}
 \end{table}
@@
   \end{tabular}
-  \label{tab:normiris}
+  \label{tab:projdigit}
 \end{table}
-Other examples ... !!!
+Table \ref{tab:normdigit} shows, for the $10$-classes digit recognition example, the norm of each Shapley composition and the cosine similarity between the, and Table \ref{tab:projdigit} shows their projection of the set of class-compositions.\\
+Add details...
 
@@
 \label{app:algo}
-In this section, we present an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we used in this work for the estimation the Shapley compositions.
+In this section, we present an adaptation of Algorithm 2 from \cite{vstrumbelj2014explaining} we used for the estimation the Shapley compositions.
 
@@
 \begin{algorithm}
-   \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
+   \caption{Adaptation of the Algorithm 1 from \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
    \label{alg:1}
@@
 \begin{algorithm}
-   \caption{Adaptation of the Algorithm 2 in \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
+   \caption{Adaptation of the Algorithm 2 from \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
    \label{alg:2}
@@
      \ENDIF
-     \STATE $\bm{\phi}_j \leftarrow \bm{\phi}_j +$result of Algorithm \ref{alg:1} for the $j$th feature and $m=1$,
+     \STATE $\bm{\phi}_j \leftarrow \bm{\phi}_j$ $+$ result of Algorithm \ref{alg:1} for the $j$th feature and $m=1$,
      \STATE update $\bm{\Sigma}^{(j)}$ using an incremental algorithm,
@@
 Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments).
-Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where is the base $\bm{f}_0$ composition, be the error composition on the pertubation of all Shapley compositions, i.e.
+Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.
 the error making the efficiency property unfulfilled.
-In order to respect the efficiency property, we want this error to be the neutral element of the perturnation, i.e.
+In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e.
 the ``zero'' in the sense of the Aitchison geometry: the uniform distribution.

### 2024-01-07 paper/main.tex
correct

@@
 \begin{abstract}
-  The Shapley value has been widely used for measuring the contribution of each feature to a model's prediction.
-However, coming from game theory, this has been designed for a one-dimensional function's codomain.
-For a multiclass probabilistic classifier, the output is a discrete probability distribution, over a set of more than two possible classes, and lives on a multidimensional simplex.
+  Shapley values have been widely used for explaining a machine learning model's prediction by measuring the contribution of each feature to the prediction.
+Coming from game theory, the Shapley value has been designed for a one-dimensional function's codomain.
+However, for a multiclass probabilistic classifier, the output is a discrete probability distribution, over a set of more than two possible classes, and lives on a multidimensional simplex.
 In this case, the Shapley values are sometimes computed on each output dimension one-by-one, in an implicit one-vs-rest setting, ignoring the compositional nature of the output distribution.
 Indeed, elements of the simplex are known as compositional data and a discrete probability distribution can therefore be treated as such taking into account the relative information between probabilities.
-Using the Aitchison geometry of the simplex, this paper presents a first initiative for a multidimensional extension of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions on the simplex in machine learning.
+Using the Aitchison geometry of the simplex, this paper presents an initiative for a multidimensional extension on the simplex of the concept of Shapley value, named Shapley composition, for explaining probabilistic predictions in machine learning.
 \end{abstract}
@@
 
-Modern machine learning approaches like the one based on deep learning are often regarded as black-boxes making them not reliable for real-life application where the machine learnbing prediction has to be understood.
-These last years, the number of contribution to make models more explainable has therefore increased in the machine learning literature.
-One way to better understand a prediction would be to measure the contribution of each input features on the computation of the model output.
-The concept of Shapley value is now widely used for this purpose \cite{vstrumbelj2014explaining,datta2016} especially since the release of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap}}.
-The Shapley value came from cooperative game theory...
+Many machine learning approaches---like the one based on deep learning or random forests---are often regarded as black-boxes making them unreliable for real-life applications where the model's predictions have to be understood or explainable.
+These recent years, the interest in more explainable models and explainability methods has therefore increased in the machine learning literature \cite{angelov2021explainable}.
+One group of approaches, known as \emph{local explanation}, aims to measure the contribution of each input feature to the computation of the model's output.
+Shapley values are widely used for this purpose \cite{vstrumbelj2014explaining,datta2016}, especially since the release of the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap}}.
+The Shapley values come from cooperative game theory where a group of players work together to maximize a payoff.
+A set of Shapley values uniquely distributes the payoff over all the players according to their individual contribution to the total.
+Shapley value is known for respecting a set of desired axiomatic properties \cite{shapley1953value}.
+For explaining a machine learning model's prediction, a player becomes a feature and the total payoff is the scalar output of the model.
 
-explain shapley in game theory,
+The concept of Shapley value is designed for a one-dimensional function's codomain.
+In game theory, the function takes a coalition of players and gives a gain or payoff.
+In machine learning, the function takes a group of features from an instance and gives a one-dimensional output or prediction.
+For two-classes classification models, the output is one-dimensional and usually given by a sigmoid transformation.
+The Shapley value framework can therefore simply be applied to the logit transformation of the model output.
+The logit transformation maps the sigmoid codomain $]0,1[$ to the real line $\mathbb{R}$.
+However, for classification tasks with more than two possible classes, the output of the model is multidimensional.
+This can be a discrete probability distribution or the output of a softmax layer commonly used in neural-networks.
+For a $N$-classes problem, these outputs live on a $(N-1)$-dimensional simplex.
+In this case, the Shapley value framework cannot be directly applied.
+Some apply the framework on each output probability one-by-one ignoring the structure of the simplex and the relative information between the probabilities.
 
-How it is applied to ML,
+Compositional data analysis \cite{aitchison1982,pawlowskymodeling} aims at treating data---known as compositions---living on a simplex.
+Compositional data analysis tools have been applied for instance on geological, and chemical data but also to discrete probability distributions \cite{egozcue2011evidence,egozcue2018evidence,noe2023representing}.
+This paper presents the use of the Aitchison geometry of the simplex, coming from the field of compositional data analysis, for extending the concept of Shapley value to the probability simplex, i.e.
+the space of discrete probability distributions.
+This can be used to properly explain multiclass probabilistic prediction.
 
-Limitation,
+The next section of the paper, Section \ref{sec:shapley}, recalls the standard definition of the Shapley values, as applied to one-dimensional predictions in machine learning.
+Section \ref{sec:compo} presents the necessary tools from compositional data analysis, in particular, the Aitchison geometry of the simplex.
+Section \ref{sec:shapcompo} is the main contribution of the paper which is mostly theoretical.
+It defines, using the Aitchison geometry, the concept of Shapley composition for extending the Shapley value framework to the multidimensional simplex.
+Section \ref{sec:explain} shows, with simple examples and visualizations, how Shapley compositions can be used for explaining multidimensional probabilistic predictions in machine learning.
+Section \ref{sec:conclud} provides additional remarks and concludes the work.
 
-...
-
-\subsection{Contributions}
-
-...
-
-\section{The Shapley value in machine learning}
+\section{The Shapley values in machine learning}
 \label{sec:shapley}
 
-In this section, we recall the theoretical formulation of the Shapley value for measuring the contribution of each feature on a machine learning prediction.
+This section recalls the theoretical formulation of the Shapley values for measuring the contribution of each feature on a machine learning prediction.
 
-Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one want to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}\subset\mathbb{R}^d$.
+Let $f:\mathcal{X}\to\mathbb{R}$ be a learned model one wants to \emph{locally} explain where $f(\bm{x})$ is the prediction on the instance $\bm{x}\in\mathcal{X}\subset\mathbb{R}^d$.
 Let $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data\footnote{Usually, this is unknown but expectations will be replaced by empirical averagings.}.
-Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$, where $d$ is the number of features that composes an instance $\bm{x}\in\mathcal{X}$, be a subset of indices.
+Let $S\subseteq \mathcal{I}=\{1,2,\dots d\}$ be a subset of indices where $d$ is the number of features.
 $\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.
@@
 When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$.
-However, when only $\bm{x}_S$ is given with $\mathcal{S} \neq \mathcal{I}$, there is uncertainty about the other features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
+However, when only $\bm{x}_S$ is given with $\mathcal{S} \neq \mathcal{I}$, there is uncertainty about the non-observed features and we therefore compute the expected prediction given $\bm{x}_S$: $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
 The contribution of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
@@
 where $2^{\mathcal{I}}$ is the set of all subsets of $\mathcal{I}$.
-This measure the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
-The whole contribution of the $i$th feature is computing by averaging this quantity over all possible coalitions as follow:
+This measures the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
+The whole contribution of the $i$th feature is computed by averaging this quantity over all possible coalitions as follows:
 \begin{equation}
@@
 where $\pi$ is a permutation of the set $\mathcal{I}$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
-For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ wil be dropt from the equations.
+For better clarity, the subscript $_{f,\bm{x},\text{Pr}}$ will be dropped from the equations.
 
@@
 
-The Shapley value is designed for one-dimensinal codomain of the function $f$.
-For explaining machine learning models which output multidimensional discrete probability distribution, like in multiclass classification, people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison.
-However, this approach ignores the relative information between each probability and ignores the compositional nature of probability distributions.
+The Shapley value is designed for a one-dimensional codomain of the function $f$.
+For explaining machine learning models which output multidimensional discrete probability distributions, like in multiclass classification, some people have been explaining each output dimension one-by-one, applying a logit transformation to the probabilities, resulting in a one-vs-rest comparison.
+However, this approach ignores the relative information between each probability and ignores the compositional nature of the discrete probability distributions.
 Indeed, the probabilistic output of a classifier lives on a multidimensional simplex.
-The latter is the sample space of data refered as \emph{compositional data} we briefly review in the next section.
+The latter is the sample space of \emph{compositional data} briefly reviewed in the next section.
 
 \section{Compositional data}
+\label{sec:compo}
 
@@
 The sample space of compositional data is the simplex: $\mathcal{S}^N = \left\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^{*N}_{+} \big| \sum_{i=1}^{N} x_i = k \right\}$.
-In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of components to handle this \cite{aitchison1982}.
+In a composition, only the relative information between parts matters and John Aitchison introduced the use of log-ratios of parts to handle this \cite{aitchison1982}.
 He defined several operations on the simplex which leads to what is called the \emph{Aitchison geometry of the simplex}.
@@
 \begin{itemize}
-\item a \emph{perturbation}:
-  $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$ seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
-\item a \emph{powering}:
-  $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$ seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
-\item an inner product:\\
-  $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$,
+\item \emph{perturbation}:
+  \begin{equation*}
+    \bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right),
+  \end{equation*}
+  seen as an addition between two compositions $\bm{x},\bm{y}\in \mathcal{S}^N$,
+\item \emph{powering}:
+  \begin{equation*}
+    \alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right),
+  \end{equation*}
+  seen as a multiplication by a scalar $\alpha \in \mathbb{R}$,
+\item inner product:
+  \begin{equation*}
+    \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j},
+  \end{equation*}
 \end{itemize}
@@
 
-This give to the simplex a $(N-1)$-dimensional Euclidean vector space structure called \emph{Aitchison geometry of the simplex}.
+This gives to the simplex a $(N-1)$-dimensional Euclidean vector space structure called \emph{Aitchison geometry of the simplex}.
 In this paper, since we are interested in classifiers' outputs as discrete probability distributions, we restrict ourselves to the \emph{probability simplex} where $k=1$.
@@
 
-An $(N-1)$-dimensional orthonormal basis of the simplex, refered as an \emph{Aitchison} orthonormal basis, can be built.
-The projection of a composition (like a discrete probability distribution) into this basis defines an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
-This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express the compositions into a Cartesian coordinates system preserving the metric of the Aitchison geometry.
-Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition, multiplication by a scalar and inner product.
+An $(N-1)$-dimensional orthonormal basis of the simplex, referred to as an \emph{Aitchison} orthonormal basis, can be built.
+The projection of a composition into this basis defines an isometric isomorphism between $\mathcal{S}^N$ and $\mathbb{R}^{N-1}$.
+This is known as an Isometric-Log-Ratio (ILR) transformation \cite{egozcue2003isometric} and allows to express a composition into a Cartesian coordinates system preserving the metric of the Aitchison geometry.
+Within this real space, the permutation, the powering and the Aitchison inner product defined above are respectively the standard addition between two vectors, the multiplication of a vector by a scalar, and the standard inner product.
 
-Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we express its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
-The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is be obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where the set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
+Given a composition $\bm{p} = \left[ p_1,\dots p_N \right]^T \in \mathcal{S}^N$ we write its ILR transformation as $\tilde{\bm{p}} = \ilr \left( \bm{p} \right) = \left[ \tilde{p}_1,\dots \tilde{p}_{N-1} \right]^T \in \mathbb{R}^{N-1}$.
+The $i$th element $\tilde{p}_i$ of $\tilde{\bm{p}}$ is obtained as: $\tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a$ where the set $\{\bm{e}^{(i)} \in \mathcal{S}^N, i=1,\dots N-1\}$ forms an \emph{Aitchison} orthonormal basis of the simplex.
 The choice of the basis will be discussed in Section \ref{sec:balances}.
 
-\section{Shapley on the simplex}
+\section{Shapley composition on the simplex}
+\label{sec:shapcompo}
 
-In Section \ref{sec:shapley} we have briefly presented the standard use of Shapley value designed for explaining one-dimensional predictions.
-In this Section, we will see how the Aitchison geometry can be used to extend the concept of Shapley value to the simplex for explaning multidimensional probabilistic predictions.
+In Section \ref{sec:shapley} we have briefly presented the standard definition of the Shapley values in the context of machine learning one-dimensional predictions explanation.
+In this Section, we will see how the Aitchison geometry can be used to extend the concept of Shapley value to the simplex for explaining multidimensional probabilistic predictions.
 
 Let $\bm{f}:\mathcal{X}\to\mathcal{S}^N$ be a learned model, like a $N$-classes probabilistic classifier for instance, which outputs a probabilistic prediction on the $(N-1)$-dimensional probability simplex $S^N$.
-To properly consider the relative information between the probabilities, the output of the model must be treated as compositional data using the operators and metric defined by the Aitchison geometry of the simplex.
-We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follow:
+In order to properly consider the structure of the simplex and the relative information between the probabilities, the output of the model must be treated as compositional data using the operators and metrics defined by the Aitchison geometry of the simplex.
+We therefore rewrite the contribution and the value function of Equations \ref{eq:contrib} and \ref{eq:valuefunction} as follows:
 \begin{equation}
@@
   \begin{aligned}
-    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{X}} &\to \mathcal{S}^N,\\
-    \bm{X}_S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
+    \bm{v}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathcal{S}^N,\\
+    S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S].
   \end{aligned}
 \end{equation}
-The $\mathcal{A}$ in superscript highlight the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
+The $\mathcal{A}$ in superscript highlights the fact that the expectation is done with respect to the Aitchison measure, rather than the Lebesgue measure, which can simply be computed as: $\mathbb{E}^{\mathcal{A}}[\bm{Y}] = \ilr^{-1}\left( \mathbb{E} \left[ \ilr\left( \bm{Y} \right) \right] \right)$,
 where $\mathbb{E}^{\mathcal{A}}$ refers to the expectation with respect to the Aitchison measure while $\mathbb{E}$ refers to the expectation with respect to the Lebesgue measure \cite{pawlowskymodeling}.
 
-The Shapley quantity expressing the contribution of the $i$th feature on a prediction can simply be expressed on the simplex as the \emph{Shapley composition} $\bm{\phi}(i)$ given by:
+The Shapley quantity expressing the contribution of the $i$th feature on a prediction can simply be expressed on the simplex as the composition $\bm{\phi}(i)$ given by:
 \begin{equation}
@@
 \end{equation}
-It can be shown (in Appendix \ref{app:properties}) that the linearity and the efficiency properties naturally hold for the Shapley composition:
+We call this quantity \emph{Shapley composition}. It can be shown (in Appendix \ref{app:properties}) that the linearity and the efficiency properties naturally hold for the Shapley composition:
 \begin{equation}
@@
 Here, the Shapley quantity is not a scalar anymore, this is a composition living on the probability simplex.
-In the next section, we will see in more details how this can be used to explain the contribution of the features on a multidimensional probabilistic prediction.
+In the next section, we will see in more detail how this can be used to explain the contribution of each feature to a multidimensional probabilistic prediction.
 
@@
 Each Shapley composition is therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}} to respect the efficiency propery.
-See Appendix \ref{app:algo} and \ref{app:correct} for more details.
+See Appendix \ref{app:algo} and \ref{app:correct} for more details\footnote{The estimation algorithm we use assumes that the features are independent.
+We leave the exploration of estimation algorithms with no features-independence assumption for future work.}.
 
-\section{Shapley values versus Shapley composition}
+% \section{Shapley values versus Shapley composition}
 
-We just proposed a theoretically well founded extension of the concept of Shapley value for the explaining multidimensional prediction in machine learning.
-The use of standard Shapley value in this context has been rarely discuss in the literature.
-However, some compute a Shapley value on each output dimension one-by-one.
-We here briefly discuss and compare this approach with the Shapley composition framework.
+\section{Discussion and conclusion}
+\label{sec:conclud}
 
-A Shapley value can be computed on the logit of the probability for each classes resuilting in $N$-dimensional vector of the Shapley values for a $N$-classes problem.
+The use of standard Shapley value for explaining machine learning model with multidimensional probabilistic output has been rarely discussed in the literature.
+However, people computing a Shapley value on each output dimension one-by-one can be encounter.
+To be more precise, one may first found natural to compute a Shapley value on the logit of the probability for each classes resuilting in a $N$-dimensional vector of the Shapley values for a $N$-classes problem.
 Such vector will be refered here as \emph{Shapley vector}.
 Even if the efficiency property holds, i.e.
-the sum of the part-wise logit of the base distribution with the Shapley vectors for each feature is equal to the part-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
+the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
+Indeed, this approach ignore the compositional nature of discrete probability distribution output by a classifier.
+Moreover, such strategy would require to run $N$ \emph{explanation model} contrary to our approach wich require a single explanation model.
 
-\section{Discussion and conclusion}
+As far as we know, this paper is the first work proposing an extension of the Shapley value framework to the multidimensional simplex for explaining probabilistic predictions in machine learning.
+We saw how the formalization of the standard Shapley value naturally extend to the simplex using the Aitchison geometry.
+The Shapley value therefore becomes a Shapley composition or distribution, i.e.
+a vector living on the probability simplex.
+Such Shapley composition informs on the contribution of a feature on the prediction.
+It tells how the feature moves the distribution from the base one to the predicted one on the simplex.
+The Aitchison geometry defines on the simplex an Euclidean vector space structure such that the Shapley compositions can be analysed through angles, norms and projections.
 
-Compare with standard Shapley ??
-
-We know small expé..., sounds tricky,... first step for a theoretically founded multiclass problems explanations...
-...
-
-Features INDEPENDENCE!!
-
-Axiomatic formulation
+The machine learning literature about the use of Shapley value is prolific.
+Many estimation algorithms have been developped, many applications of the Shapley value have emerged, and large scales experiments have been conducted.
+In contrast, our paper presents limited experimental results as simple proofs of concept and illustrations.
+However, we believe this work lies down proper theoretical foundations to foster researches in explainable machine learning especially for multiclass classification.
 

### 2024-01-07 paper/main.tex
correct proof read

@@
 Compositional data analysis tools have been applied for instance on geological, and chemical data but also to discrete probability distributions \cite{egozcue2011evidence,egozcue2018evidence,noe2023representing}.
-This paper presents the use of the Aitchison geometry of the simplex, coming from the field of compositional data analysis, for extending the concept of Shapley value to the probability simplex, i.e.
-the space of discrete probability distributions.
+This paper presents the use of the Aitchison geometry of the simplex, coming from the field of compositional data analysis, for extending the concept of Shapley value to the probability simplex, i.e.~the space of discrete probability distributions.
 This can be used to properly explain multiclass probabilistic prediction.
@@
 
-\section{Explaining probabilistic prediction with Shapley compositions}
+\section{Explaining a prediction with Shapley compositions}
 \label{sec:explain}
 
-Given a prediction $\bm{f}(\bm{x})$, the Shapley composition $\bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i)$ describes the contribution of the $i$th feature on the prediction.
-The efficiency property shows how the probability distribution moves from the base value, i.e.
-the expected prediction regardless of the current input, to the prediction $\bm{f}(\bm{x})$.
+Given a probabilistic prediction $\bm{f}(\bm{x}) \in \mathcal{S}^N$, the Shapley composition $\bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i)$ describes the contribution of the $i$th feature on the prediction.
+The efficiency property shows how the probability distribution moves from the base distribution, i.e.~the expected prediction regardless of the current input, to the prediction $\bm{f}(\bm{x})$.
 In the standard Shapley formulation recalled in Section \ref{sec:shapley}, the prediction is one-dimensional such that the Shapley quantity is a scalar.
-In application where there are more than two possible classes, the prediction is multidimensional such that the Shapley quantity is too.
-Both lives on the same space: the probability simplex.
-In this section, we discuss how the set of Shapley compositions can be analysed to better understand the contribution and influence of each features on the prediction.
+In applications where there are more than two possible classes, the prediction is multidimensional such that the Shapley quantity is too.
+Both live in the same space: the probability simplex.
+In this section, we discuss how the set of Shapley compositions can be analysed to better understand the contribution and influence of each feature on the prediction.
 
@@
 The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}.
-This space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.
+As we will see, this space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.
 
@@
 
-In the three classes case, the space is $2$-dimensional.
-We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width.
-The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to.
-In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier.
-Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction.
-Having the highest norm, the petal length is the feature contributing the most on the prediction, and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines).
-Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class.
-The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class.
-The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
+In a three-classes case, the space of the predictions is $2$-dimensional.
+We illustrate this example with the well-known Iris classification dataset consisting of a set of flowers described by 4 features: sepal length and width, and petal length and width.
+The aim of the classification task is to predict to which of the three species (setosa, versicolor and virginica) a flower belongs.
+In the present example, a Support Vector Machine (SVM) with a radial basis function (rbf) kernel and with pairwise coupling \cite{wu2003probability} is used as a probabilistic classifier.
+Figure \ref{fig:3classes} shows the explanation of the classifier prediction with one versicolor instance as input.
+Figure \ref{fig:3classesshap} shows the Shapley compositions in the ILR space and Figure \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction.
+Having the highest norm, the petal length is the feature contributing the most to the prediction and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines).
+Class-compositions are represented by color dashed vectors.
+A class-composition is defined as an unit norm composition going straight to the direction of one class and against all the others (see Appendix \ref{app:classcompo} for a formal definition).
+The Shapley composition for the petal length is orthogonal to the virginica class-composition, this suggests that this feature does not contribute to the predicted probability for this class.
+The Shapley composition for the petal width goes straight to the opposite direction of the setosa class-composition suggesting that this feature contributes to rejecting this class.
+The other Shapley compositions have a low norm suggesting that these features do not contribute to the prediction.
 \begin{figure}
-
   \begin{subfigure}{0.5\textwidth}
@@
     \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
-    \caption{Sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
+    \caption{The sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
     \label{fig:3classesshapsum}
   \end{subfigure}
-
   \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
@@
 
-In a four classes example, the simplex is $3$-dimensional.
-We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}.
-It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3.
-Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley composition.
+In a four-classes example, the simplex is $3$-dimensional.
+We illustrate this with a simple handwritten digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}.
+It consists of classifying an $8\times8$ image as representing one of the digits among 0, 1, 2 and 3.
+Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley compositions.
+This would be hard to visualize and analyse.
 We therefore reduce the number of features to 6 using a principal component analysis for better clarity and conciseness.
-We again use a SVM classifier with a rbf kernel.
-The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}.
-To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule.
-The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e.
-a $3$-dimensional simplex.
-The relative position of the class-compositions in the ILR space are the same as the bonds between the carbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
-In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double blind peer reviewing.}.
+A SVM with a rbf kernel and pairwise coupling is again used as a probabilistic classifier.
+The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated in Figure \ref{fig:4classesshapsum}.
+To better understand how this space is divided into four regions---each representing the maximum probability region for one class---one can think about the shape of a methane molecule.
+The hydrogens correspond to the vertices and the carbon to the center of a tetrahedron i.e.~a $3$-dimensional simplex.
+The relative positions of the class-compositions in the ILR space are the same as the bonds between the carbon and hydrogen: the angles are $\approx 109.5^{\circ}$.
+In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double-blind peer reviewing.}.
 \begin{figure}
@@
 
-When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once, but $2$ or $3$-dimensional subspaces can still be visualized.
-In order to select the ILR components to visualize, one needs to understand what they refer to.
+When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once.
+However, $2$ or $3$-dimensional subspaces can still be visualized.
+In order to select the ILR components to investigate, one needs to understand what they refer to.
 In this section, we briefly discuss the interpretation of the ILR components.% of the explanation of a classifier prediction with more than three classes.
 
-A component of the ILR space  can be interpreted as a \emph{balance}, i.e.
-a log-ratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes.
+A component of the ILR space can be interpreted as a \emph{balance}, i.e.~a log-ratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes.
 Therefore, a balance is here comparing the weight of two groups of classes.
@@
 Figure \ref{fig:trees} gives two examples: \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$.
-The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$.
-Each next balance then recursively compares the probability for the next class with the probabilities for the previous ones independently of all the others.
+Each node of the tree is a balance i.e.~an ILR component.
+The first balance $\tilde{p}_1$ first compares the probability for class $1$ with the probability for class $2$.
+Each next balance then recursively compares the probability for the next class with the probabilities for the previous classes independently of all the others.
 
-In some applications, one may be interested in particular comparisons of groups of classes.
-For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups.
+In some applications, one may be interested in particular comparisons of groups of classes not necessarily given by a basis in the form of Figure \ref{fig:bifurc1}.
+For instance, as in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups.
 However, there are sometimes no obvious relevant comparisons to study.
-For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent.
+For instance, in the handwritten digit recognition problem, one may want to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent.
 Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy.
-Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}.
+
+Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition task\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging a pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with the same covariance matrix.}.
 We comment, for conciseness, a single $2$-dimensional subspace\footnote{More visualisations can be obtaned from the shared python notebooks.}.
 Let's have a look, in Figure \ref{fig:moreclasses35}, at the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$).
-It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8.
+It is like saying we are only interested in comparing the probability assigned for class 0 with the probability assigned for class 6, and in comparing the probability assigned for class 1 with the group of probabilities assigned for classes 7 and 8.
 $\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8.
-Therefore, the class-composition for the others digits have a zero projection within this subspace and are not drawn in Figure \ref{fig:moreclasses35}.
-The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8.
-Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor.
+Therefore, the class-compositions for the other digits have a zero projection within this subspace and are therefore discarded in Figure \ref{fig:moreclasses35}.
+The class-compositions for 0 and 6 are orthogonal to the class-compositions for classes 1, 7 and 8.
+Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection.
+In contrast, in the example of Figure \ref{fig:3classes}, $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor.
 In this case, none of the class-compositions are orthogonal.
-In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection of the class-compositions on this line for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8.
-The two latters, are equal and are half as long as the former.
+In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection on this line of the class-compositions for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8.
+The latter two are equal and half as long as the former.
 In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.
+In other words, in this subspace, the class-compositions for 7 and 8 are reweighted such that this group of two classes has the same weight as the group made of the single class 1.
 
@@
       \input{bifurctree1.tex}
-      \caption{Bifurcation tree used in our $10$-classes digit recognition task}
+      \caption{Bifurcation tree used in the $10$-classes digit recognition task discussed in Section \ref{sec:balances} and in Figure \ref{fig:moreclasses35}.}
       \label{fig:bifurc2}
@@
   \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
-  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from the class 2.
-$\tilde{p}_3$ compares the probability for the class 0 with the one for the class 6 and $\tilde{p}_5$ compares the probability for the class 1 with the group of probabilities for the class 7 and 8.
-}
+  \caption{The sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from class 2.
+$\tilde{p}_3$ compares the probability assigned for class 0 with the probability assigned for class 6 and $\tilde{p}_5$ compares the probability assigned for class 1 with the group of probabilities assigned for class 7 and 8.
+The color dashed vectors represent the class-compositions with non-zero projection.}
   \label{fig:moreclasses35}
@@
 Such vector will be refered here as \emph{Shapley vector}.
-Even if the efficiency property holds, i.e.
-the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
+Even if the efficiency property holds, i.e.~the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
 Indeed, this approach ignore the compositional nature of discrete probability distribution output by a classifier.
@@
 We saw how the formalization of the standard Shapley value naturally extend to the simplex using the Aitchison geometry.
-The Shapley value therefore becomes a Shapley composition or distribution, i.e.
-a vector living on the probability simplex.
+The Shapley value therefore becomes a Shapley composition or distribution, i.e.~a vector living on the probability simplex.
 Such Shapley composition informs on the contribution of a feature on the prediction.
@@
 \section{Class-compositions}
-
+\label{app:classcompo}
 A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
@@
 Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments).
-Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.
-the error making the efficiency property unfulfilled.
-In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e.
-the ``zero'' in the sense of the Aitchison geometry: the uniform distribution.
-We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.
-with a precision likely to be lower) to move more than the ones with a smaller variance (i.e.
-with a precision likely to be higher).
+Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.~the error making the efficiency property unfulfilled.
+In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e.~the ``zero'' in the sense of the Aitchison geometry: the uniform distribution.
+We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.~with a precision likely to be lower) to move more than the ones with a smaller variance (i.e.~with a precision likely to be higher).
 

### 2024-01-08 paper/main.tex
1st version

@@
 Class-compositions are represented by color dashed vectors.
-A class-composition is defined as an unit norm composition going straight to the direction of one class and against all the others (see Appendix \ref{app:classcompo} for a formal definition).
+A class-composition is defined as an unit norm composition going straight to the direction of one class and uniformly against all the others (see Appendix \ref{app:classcompo} for a formal definition).
 The Shapley composition for the petal length is orthogonal to the virginica class-composition, this suggests that this feature does not contribute to the predicted probability for this class.
@@
 
-Some may find the analysis of the features contributions in cases with more than four classes tricky.
-Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the balances and a careful building of the bifurcation tree.
-However, the Shapley explaination can be summarized by sets of angles, norms and projections.
-The norm of a Shapley composition gives the strength of the feature's contribution in the prediction.
-It gives the overall contribution of the feature on the prediction, regardless of its direction.
-The angle between two Shapley compositions can informs about their orthogonality.
-If two Shapley composition are orthogonal%\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}
-, this suggest the features are nonredundant.
-A negative angle would suggest that the features have an opposite influence on the prediction.
-The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing.
-Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.
+Some may find the analysis of the contribution of each feature using Shapley compositions in cases with more than four classes tricky.
+Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize requires a careful understanding of the balances, the Aitchison basis and the bifurcation tree.
+However, the Shapley explanation can be summarized by sets of angles, norms and projections:
+\begin{itemize}
+\item The norm of a Shapley composition gives the strength of the feature's contribution to the prediction.
+The measures the overall contribution of the feature on the prediction, regardless of its direction.
+\item The angle between two Shapley compositions can inform about their orthogonality.
+If two Shapley compositions are orthogonal%\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}
+, this suggests that the features are nonredundant. A negative angle would suggest that the features have an opposite influence on the prediction.
+\item The projections of a Shapley composition on the set of class-compositions inform in favor of, or against, which classes a feature is contributing.
+\end{itemize}
+
+  Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, the angles between them and their projections on the class-compositions.
 
@@
 
-If one found hard to visualize the proposed Shapley explanation in the ILR space, the Shapley composition can be visualized as histograms like discrete probability distributions.
-Figure \ref{fig:histiris} shows the Shapley compositions of the Iris classification example.
-The more uniform the histogram is, like for the sepal length and width, the less the contribution of the feature is.
-In opposite, the histogram for the petal length as a high value for the versicolor class, relatively to the others, confirming the contribution of the feature toward this class.
-For the petal width, the value for the class setosa is low relatively to the others which confirms the constribution of this feature against the class setosa.
+If one found hard to visualize the proposed Shapley explanation in the ILR space, the Shapley composition can be visualized using a bar plot like discrete probability distributions.
+Figure \ref{fig:histiris} shows, as histograms, the Shapley compositions of the Iris classification example.
+The more uniform a histogram is, like for the sepal length and width, the less the contribution of the feature is.
+In contrast, the histogram for the petal length has a high value for the versicolor class, in comparison to the others.
+This confirms the contribution of the feature toward this class.
+For the petal width, the value for the class setosa is low in comparison to the others which confirms the contribution of this feature against the class setosa.
 \begin{figure}
@@
 \end{figure}
-
 As another illustration, Figure \ref{fig:histmore} shows the Shapley compositions of the $10$-classes digit recognition example.
-Contrary to the visualization of the compositions within the ILR space as discussed in Section \ref{sec:balances}, here, one can analyses all parts of each compositions within a single plot.
-In this example, the high of the class 2 for the first principal component confirms the contribution of this features toward this class.
+Here, and contrary to the visualization of the compositions within the ILR space as discussed in Section \ref{sec:balances}, one can analyze all parts of each composition within a single plot.
+In this example, the high value for class 2 for the first principal component confirms the contribution of this feature toward this class.
 \begin{figure}
@@
 
-In this work, the estimation algorithm we used to compute the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}.
+The estimation algorithm used in this work for computing the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}.
 Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold.
-Each Shapley composition is therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}} to respect the efficiency propery.
-See Appendix \ref{app:algo} and \ref{app:correct} for more details\footnote{The estimation algorithm we use assumes that the features are independent.
-We leave the exploration of estimation algorithms with no features-independence assumption for future work.}.
-
-% \section{Shapley values versus Shapley composition}
+In order for the set of estimated Shapley compositions to respect the efficiency property, each Shapley composition is adjusted following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
+See Appendix \ref{app:algo} and \ref{app:correct} for more details\footnote{The estimation algorithm assumes that the features are independent.
+The estimation of the Shapley values without such assumption has been discussed in the literature \cite{AAS2021103502}.
+We leave the exploration of estimation algorithms of the Shapley compositions without features-independence assumption for future work.}.
 
@@
 
-The use of standard Shapley value for explaining machine learning model with multidimensional probabilistic output has been rarely discussed in the literature.
-However, people computing a Shapley value on each output dimension one-by-one can be encounter.
-To be more precise, one may first found natural to compute a Shapley value on the logit of the probability for each classes resuilting in a $N$-dimensional vector of the Shapley values for a $N$-classes problem.
-Such vector will be refered here as \emph{Shapley vector}.
-Even if the efficiency property holds, i.e.~the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive.
-Indeed, this approach ignore the compositional nature of discrete probability distribution output by a classifier.
-Moreover, such strategy would require to run $N$ \emph{explanation model} contrary to our approach wich require a single explanation model.
+The use of standard Shapley value for explaining machine learning models with multidimensional probabilistic output has been rarely discussed in the literature.
+However, people computing a Shapley value on each output dimension one-by-one can be encountered.
+To be more precise, for a $N$-classes problem ($N>2$), one may first found natural to compute a Shapley value on the logit of the probability for each class resulting in a $N$-dimensional vector of the Shapley values we call \emph{Shapley vector}.
+Even if the efficiency property holds with the standard addition, i.e.~the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is counterintuitive.
+Indeed, this is because this approach ignores the compositional nature of discrete probability distribution output by a probabilistic classifier.
+Moreover, such a strategy would require running $N$ \emph{explanation models} contrary to the Shapley composition approach which requires a single explanation model.
 
 As far as we know, this paper is the first work proposing an extension of the Shapley value framework to the multidimensional simplex for explaining probabilistic predictions in machine learning.
-We saw how the formalization of the standard Shapley value naturally extend to the simplex using the Aitchison geometry.
-The Shapley value therefore becomes a Shapley composition or distribution, i.e.~a vector living on the probability simplex.
+We saw how the formalization of the standard Shapley value naturally extends to the simplex using the Aitchison geometry coming from compositional data analysis.
+The Shapley quantity therefore becomes a composition or distribution, i.e.~a vector living on the probability simplex.
 Such Shapley composition informs on the contribution of a feature on the prediction.
 It tells how the feature moves the distribution from the base one to the predicted one on the simplex.
-The Aitchison geometry defines on the simplex an Euclidean vector space structure such that the Shapley compositions can be analysed through angles, norms and projections.
+The Aitchison geometry gives to the simplex an Euclidean vector space structure such that the Shapley compositions can be analysed through angles, norms and projections.
 
 The machine learning literature about the use of Shapley value is prolific.
-Many estimation algorithms have been developped, many applications of the Shapley value have emerged, and large scales experiments have been conducted.
-In contrast, our paper presents limited experimental results as simple proofs of concept and illustrations.
-However, we believe this work lies down proper theoretical foundations to foster researches in explainable machine learning especially for multiclass classification.
+Many estimation algorithms have been developed, many applications of the Shapley value have emerged, and large-scale experiments have been conducted.
+In contrast, our paper presents limited experimental results as simple proofs of concepts and illustrations.
+However, we believe this work lies down proper theoretical foundations to foster the research in explainable machine learning, especially for multiclass recognition problems.
 
@@
 
-The Shapley composition is linear, within the Aitchison geometry of the simplex, with respect to linear combination of models' predictions.
+The Shapley composition is linear, within the Aitchison geometry of the simplex, with respect to linear combinations of models' prediction.
 \begin{proof}
-  Let's consider the linear combination of predictions $\bm{h}(\bm{x}) = \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})$. we want to check if:
+  Let's consider the linear combination of predictions $\bm{h}(\bm{x}) = \alpha \odot \bm{f}(\bm{x}) \oplus \beta \odot \bm{g}(\bm{x})$. We want to check if:
   \begin{equation}
@@
   \end{equation}
-  And the linearity of the Shap composition:
+  And the linearity of the Shapley composition:
   \begin{equation}
@@
 \label{app:classcompo}
-A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
+A $k$-class-composition $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class.
 This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others.
-The $i$th part of the $\bm{c}^{(k)}$ is:
+The $i$th part of $\bm{c}^{(k)}$ is:
 \begin{equation}
@@
 \end{equation}
-where $p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
+where $0<p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
 \begin{equation}
@@
     \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
-    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
+    \text{since $p<\frac{1}{N}$ and the norm is positive:}&\\
     \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
@@
 \end{equation}
-To summarize, the $i$th part of a $k$-class-compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
+To summarize, the $i$th part of a $k$-class-composition $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
 \begin{equation}
@@
   \centering
-  \caption{Norm of the Shapley compositions, projection on the class-compositions and cosine similarity betweeen the Shapley compositions for the Iris classification example.}
+  \caption{Norm of the Shapley compositions, projections of the Shapley compositions on the class-compositions, and cosine similarity betweeen the Shapley compositions for the Iris classification example.}
   \begin{tabular}{c|c|ccc|cccc|}
@@
 \end{table}
-The Table \ref{tab:normiris} gives, for the Iris classification example of Figure \ref{fig:3classes}, the norm of each Shapley compositions, the cosine similarity between them and their projection on the set of class-compositions.
-Having the highest norm, this confirms that the petal length is the feature that contributed the most on the prediction.
-Having a close to zero inner product with the virginica class-composition, shows that this feature did not contribute to the value of the probability of of this class.
-It contributes neither in favor, nor in the reject of this class.
-Having a positive inner product with the versicolor class-composition and a negative one with the setosa class-composition, this suggest that this feature is going in favor of the class versicolor and against setosa.
-
 \begin{table}
   \centering
-  \caption{Norm of the Shapley compositions and Cosine similarity betweeen the Shapley composition for the $10$-classes digit recognition example.}
+  \caption{Norm of the Shapley compositions and Cosine similarity betweeen the Shapley compositions for the $10$-classes digit recognition example.}
   \begin{tabular}{c|c|cccccc|}
@@
 \end{table}
-
 \begin{table}
@@
 \end{table}
-Table \ref{tab:normdigit} shows, for the $10$-classes digit recognition example, the norm of each Shapley composition and the cosine similarity between the, and Table \ref{tab:projdigit} shows their projection of the set of class-compositions.\\
-Add details...
+The Table \ref{tab:normiris} gives, for the Iris classification example of Figure \ref{fig:3classes}, the norm of each Shapley composition, the cosine similarity between them and their projections on the set of class-compositions.
+With the highest norm, this confirms that the petal length is the feature that contributes the most to the prediction.
+Having a close to zero inner product with the virginica class-composition, shows that this feature does not contribute to the value of the probability assigned for this class.
+This feature contributes neither in favor, nor in the rejection of this class.
+With a positive inner product with the versicolor class-composition and a negative one with the setosa class-composition, this suggests that this feature is going in favor of the class versicolor and against the class setosa.
+
+Table \ref{tab:normdigit} shows, for the $10$-classes digit recognition example, the norm of each Shapley composition and the cosine similarity between them.
+Table \ref{tab:projdigit} shows their projections on the set of class-compositions.
 
@@
 \label{app:algo}
-In this section, we present an adaptation of Algorithm 2 from \cite{vstrumbelj2014explaining} we used for the estimation the Shapley compositions.
+This section presents an adaptation of Algorithm 2 from \cite{vstrumbelj2014explaining} we used for the estimation the Shapley compositions.
 
@@
      \item $\bm{b}_1$: which takes the values from $\bm{x}$ for the $i$th feature and the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise,
-     \item $\bm{b}_2$: which takes the values from $\bm{x}$ the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise.
+     \item $\bm{b}_2$: which takes the values from $\bm{x}$ for the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise.
      \end{itemize}
@@
 \begin{algorithm}
-   \caption{Adaptation of the Algorithm 2 from \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
+   \caption{Adaptation of the Algorithm 2 from \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples for each feature estimation.}
    \label{alg:2}
@@
 This point is often overlooked in the literature.
-In this section we write down an adjustment strategy of the estimated Shapley compositions for them to respect the efficiency property.
+In this section, we write down an adjustment strategy of the estimated Shapley compositions for them to respect the efficiency property.
 This is a similar strategy as in the sampling approximation of the Shapley values in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}.
@@
 Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments).
-Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.~the error making the efficiency property unfulfilled.
+Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.~the error that makes the efficiency property unfulfilled.
 In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e.~the ``zero'' in the sense of the Aitchison geometry: the uniform distribution.
-We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.~with a precision likely to be lower) to move more than the ones with a smaller variance (i.e.~with a precision likely to be higher).
+We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the Shapley compositions with a higher estimation variance (i.e.~with a precision likely to be lower) to move more than the Shapley compositions with a smaller estimation variance (i.e.~with a precision likely to be higher).
 
-We therefore weight the $i$th adjustment by a scalar $w_i = w\left(\tr\left(\bm{\Sigma}^{(i)}\right)\right)$, where $w$ is an increasing function, and where $\displaystyle \sum_{i=1}^d w_i= 1$.
-Note that the vector of weight is actually a composition too.
+The $i$th adjustment is therefore weighted by a scalar $w_i = w\left(\tr\left(\bm{\Sigma}^{(i)}\right)\right)$, where $w$ is an increasing function, and where $\displaystyle \sum_{i=1}^d w_i= 1$ (note that the vector of weight is actually a composition too).
 Similarly to the SHAP toolkit implementation, we choose $w$ as:
@@
 \end{equation}
-In this way, when $\epsilon$ goes to zero\footnote{In our experiments, $\epsilon = 10^{-6}$.}, the efficiency property is respected for the adjusted Shapley compositions and more weight is given to the adjusments of the Shapley compositions with a higher estimation variance.
+In this way, when $\epsilon$ goes to zero\footnote{In our experiments, $\epsilon = 10^{-6}$.}, the efficiency property is respected for the adjusted Shapley compositions and more weight is given to the adjustments of the Shapley compositions with a higher estimation variance.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

### 2024-01-16 presentation/main.tex
presentation

\documentclass{beamer}
\usetheme{Boadilla}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,chains}
\usetikzlibrary[calc]
\usepackage{amssymb,amsmath,mathrsfs,stmaryrd,bm,amsthm,amsfonts}
\usepackage{pgfplotstable}
\usepackage{biblatex}
\addbibresource{../paper/biblio.bib}
\renewcommand{\footnotesize}{\scriptsize}
\usepackage{pifont}
\usetikzlibrary{math}

\newcommand{\tikzxmark}{
\tikz[scale=0.25] {
    \draw[line width=0.7,color=red,line cap=round] (0,0) to [bend left=6] (1,1);
    \draw[line width=0.7,color=red,line cap=round] (0.2,0.95) to [bend right=3] (0.8,0.05);
}}
\newcommand{\tikzcmark}{
\tikz[scale=0.25] {
    \draw[line width=0.7,color=green,line cap=round] (0.25,0) to [bend left=10] (1,1);
    \draw[line width=0.8,color=green,line cap=round] (0,0.35) to [bend right=1] (0.23,0);
}}

%Information to be included in the title page:
\title[Shapley compositions]{Explaining probabilistic predictions on the simplex\\with Shapley compositions}
\author[Paul-Gauthier Noé]{\textbf{Paul-Gauthier Noé} \inst{1}, Miquel Perelló-Nieto \inst{2},\\Peter Flach \inst{2}, Jean-François Bonastre \inst{1}}
\institute[]{\inst{1} Avignon Université \and \inst{2} University of Bristol}

\date[LIA seminar]
{LIA seminar, Avignon Université\\January 18, 2024}
\begin{document}

\frame{\titlepage}

\section{Introduction}

\begin{frame}
\frametitle{Introduction}

\textbf{Local explanation in machine learning:}

\begin{figure}
  \centering
  \begin{tikzpicture}
  \node [draw,
	minimum width=2cm,
	minimum height=1.2cm
        ]  (model) {Model};
        \node [left=1cm of model] (x) {$\bm{x}\in\mathbb{R}^d$};
        \node [right=1cm of model] (pred) {Prediction};
        \draw[-stealth] (x.east) -- (model.west);
        \draw[-stealth] (model.east) -- (pred.west);
      \end{tikzpicture}
    \end{figure}
    \textit{Given one instance $\bm{x}$ with $d$ features, what is the contribution/effect of a feature's value on the prediction?}

    \begin{center}
      $\neq$ Global explanation
    \end{center}
  \end{frame}

  \begin{frame}
    \frametitle{Introduction}
    Examples of local explanation methods:
    \begin{itemize}
    \item Local Interpretable Model-Agnostic Explanations (LIME) \footfullcite{ribeiro2016should},
    \item Shapley values \footfullcite{vstrumbelj2014explaining} (SHAP toolkit \footfullcite{NIPS2017_7062})

    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Introduction}
    \textbf{Shapley values in cooperatibe game theory}\footfullcite{shapley1953value}
    \begin{itemize}
    \item Distributes the total payoff among the players.
    \item The unique quantity respecting a set of desired axiomatic properties:
      \begin{itemize}
      \item Linearity:\\
        \begin{equation}
          \phi_{\alpha v + (1-\alpha)w}(i) = \alpha \phi_{v}(i) + (1-\alpha) \phi_{w}(i),
        \end{equation}
        for a player $i$ and two games $v$ and $w$, and for $\alpha \in [0,1]$
      \item Efficiency,
        \begin{equation}
          \sum_{i\in\mathcal{C}} \phi_{v}(i) = v (\mathcal{C}),
        \end{equation}
        (the sum of the value is equal to the total payoff)
      \item Symmetry
      \end{itemize}
    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Introduction}
    \textbf{Shapley values in machine learning}
    \begin{itemize}
    \pause
    \item Features are treated as players, and the scalar output of the model as the payoff,
    \begin{figure}
  \centering
  \begin{tikzpicture}
  \node [draw,
	minimum width=2cm,
	minimum height=1.2cm
        ]  (model) {Model};
        \node [left=1cm of model] (x) {$\bm{x}\in\mathbb{R}^d$};
        \node [right=1cm of model] (pred) {\begin{tabular}{c}$y\in\mathbb{R}$\\\scriptsize scalar output\end{tabular}};
        \draw[-stealth] (x.east) -- (model.west);
        \draw[-stealth] (model.east) -- (pred.west);
      \end{tikzpicture}
    \end{figure}
    \pause
    \item Binary classifier, regressor with one-dimensional output \tikzcmark
    \pause
    \item Multiclass classifier \tikzxmark

          ex: The output of a softmax layer lives on a multidimensional simplex!
        \end{itemize}

  \end{frame}

  \begin{frame}
    \frametitle{Introduction}

\tikzset{%
  every neuron/.style={
    circle,
    draw,
    minimum size=0.5cm
  },
  neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}

\begin{figure}
  \centering
  \begin{tikzpicture}[>=stealth]

\foreach \m/\l [count=\y] in {1,2,3,missing,4}
  \node [every neuron/.try, neuron \m/.try] (input-\m) at (0,2.5-\y) {};

\foreach \m [count=\y] in {1,missing,2}
  \node [every neuron/.try, neuron \m/.try ] (hidden-\m) at (2,2-\y*1.25) {};

\foreach \m [count=\y] in {1,2,missing,3}
  \node [every neuron/.try, neuron \m/.try ] (output-\m) at (4,2-\y) {};

\foreach \l [count=\i] in {1,2,3,n}
  \draw [<-] (input-\i) -- ++(-1,0)
    node [above, midway] {$x_\l$};

\foreach \l [count=\i] in {1,n}
  \node [above] at (hidden-\i.north) {};

\foreach \l [count=\i] in {1,2,N}
  \draw [->] (output-\i) -- ++(1,0)
  node [above, midway] {$p_\l$};

\foreach \l [count=\i] in {1,2,N}
  \draw [very thick, color=red] (output-\i)+(1,0) circle (0.2);

\foreach \i in {1,...,4}
  \foreach \j in {1,...,2}
    \draw [->] (input-\i) -- (hidden-\j);

\foreach \i in {1,...,2}
  \foreach \j in {1,...,3}
    \draw [->] (hidden-\i) -- (output-\j);

  \end{tikzpicture}
\end{figure}
\begin{center}
  Some explain the output one-by-one,
\end{center}
\pause

        \begin{center}
          But a probability distribution lives on a simplex,\\
          \textbf{The relative information matter!!}
        \end{center}

      \end{frame}

  \begin{frame}
    \frametitle{Introduction}

\tikzset{%
  every neuron/.style={
    circle,
    draw,
    minimum size=0.5cm
  },
  neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}

\begin{figure}
  \centering
  \begin{tikzpicture}[>=stealth]

\foreach \m/\l [count=\y] in {1,2,3,missing,4}
  \node [every neuron/.try, neuron \m/.try] (input-\m) at (0,2.5-\y) {};

\foreach \m [count=\y] in {1,missing,2}
  \node [every neuron/.try, neuron \m/.try ] (hidden-\m) at (2,2-\y*1.25) {};

\foreach \m [count=\y] in {1,2,missing,3}
  \node [every neuron/.try, neuron \m/.try ] (output-\m) at (4,2-\y) {};

\foreach \l [count=\i] in {1,2,3,n}
  \draw [<-] (input-\i) -- ++(-1,0)
    node [above, midway] {$x_\l$};

\foreach \l [count=\i] in {1,n}
  \node [above] at (hidden-\i.north) {};

\foreach \l [count=\i] in {1,2,N}
  \draw [->] (output-\i) -- ++(1,0)
  node [above, midway] {$p_\l$};

\foreach \i in {1,...,4}
  \foreach \j in {1,...,2}
    \draw [->] (input-\i) -- (hidden-\j);

\foreach \i in {1,...,2}
  \foreach \j in {1,...,3}
    \draw [->] (hidden-\i) -- (output-\j);

  \draw [very thick, color=red] (5,-0.5) ellipse (0.2 and 2);

  \end{tikzpicture}
\end{figure}
        \begin{center}
          We will explain the probablities all together using the\\
          \emph{Aitchison geometry of the simplex}\footfullcite{aitchison1982,pawlowskymodeling}.
        \end{center}

  \end{frame}

  \begin{frame}{Outline}
    \tableofcontents
  \end{frame}

\section{The Shapley values in machine learning}

\begin{frame}
\frametitle{The Shapley values in machine learning}

We want to explain a prediction $f(\bm{x})$ on the instance $\bm{x}\in\mathcal{X}\subset\mathbb{R}^d$, where $f:\mathcal{X}\to\mathbb{R}$ is the learned model.
\begin{figure}
  \centering
  \begin{tikzpicture}
  \node [draw,
	minimum width=1cm,
	minimum height=1cm
        ]  (model) {$f$};
        \node [left=1cm of model] (x) {$\bm{x}$};
        \node [right=1cm of model] (pred) {\begin{tabular}{c}$f(\bm{x})\in\mathbb{R}$\\\scriptsize prediction\end{tabular}};
        \draw[-stealth] (x.east) -- (model.west);
        \draw[-stealth] (model.east) -- (pred.west);
      \end{tikzpicture}
    \end{figure}

\pause
Some notation. Let:
\begin{itemize}
\item $\text{Pr}$ be the probability distribution over $\mathcal{X}$ of the data.
\pause
\item $S\subseteq \mathcal{I}=\{1,2,\dots d\}$ be a subset of indices,
\pause
\item $\bm{x}_S$ refers to an instance $\bm{x}$ restricted to the features indicated by the indices in $S$.
\end{itemize}

\end{frame}

\begin{frame}
  \frametitle{The Shapley values in machine learning}
The \textbf{value function}:
\begin{equation}
  \begin{aligned}
    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
  \end{aligned}
\end{equation}
where $\mathbb{E}_{\text{Pr}}[f(\bm{x}) \mid \bm{x}_S] = \int_{\bm{x} \in \mathcal{X}}f(\bm{x})\text{Pr}(\bm{x} \mid \bm{x}_S)d\bm{x}$.
\vspace{1cm}

When an instance $\bm{x}$ is observed, the expected value of the prediction is simply $\mathbb{E}[f(\bm{x}) \mid \bm{x}] = f(\bm{x})$. However, when only $\bm{x}_S$ is given with $\mathcal{S} \neq \mathcal{I}$, there is uncertainty about the non-observed features and we therefore compute the expected prediction given $\bm{x}_S$.
\end{frame}

\begin{frame}
\frametitle{The Shapley values in machine learning}

The \textbf{contribution} of the feature indexed by $i \notin S$ in the prediction $f(\bm{x})$ given the known features indexed by $S$ is given by:
\begin{equation}
  \label{eq:contrib}
  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
\end{equation}

This measures the contribution of the $i$th features with a particular \emph{coalition} of features indexed by $S$.
\end{frame}

\begin{frame}
  \frametitle{The Shapley values in machine learning}

 The whole contribution of the $i$th feature is computed by averaging this quantity over all possible coalitions of features as follows:
\begin{equation}
  \phi_{f,\bm{x},\text{Pr}}(i) = \frac{1}{d!} \sum_{\pi}c_{f,\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}),
\end{equation}
where $\pi$ is a permutation of the set $\mathcal{I}$ of indexes and $\pi^{<i}_{\bm{X}}$ is the features of $\bm{X}$ coming before the $i$th feature in the ordering given by $\pi$.
\pause
\begin{center}
  This quantity is known as the {\Large\textbf{Shapley value}} for the $i$th feature.
\end{center}

\end{frame}

\begin{frame}
  \frametitle{The Shapley values in machine learning}

  It comes from cooperative game theory and is known to be the only quantity respecting a set of desired axiomatic properties \footfullcite{shapley1953value}.
  \begin{itemize}
    \pause
\item Linearity with respect to the model ($\alpha, \beta \in \mathbb{R}$): $\phi_{\alpha f +\beta g}(i) = \alpha \phi_f(i) + \beta \phi_g(i)$,
    \pause
\item The ``centered'' learned model is additively separable with respect to the Shapley values:
\begin{equation}
  f(\bm{x})-\mathbb{E}_{\text{Pr}}[f(\bm{X})] = \sum_{i=1}^{d} \phi_f(i),
\end{equation}
which is known as the \emph{efficiency} property.
    \pause
\item Symmetry
\end{itemize}

\end{frame}

\begin{frame}
  \frametitle{The Shapley values in machine learning}
  Example of explanation:
  \begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/example_from_shap}
    \caption{Explanation of the probability for the class Setosa for a flower from the Iris dataset. The classifier is an SVM with radial basis function and pairwise coupling. {\tiny Image from \url{https://github.com/shap/shap/tree/master}.}}
  \end{figure}

  Note that the Shapley explanation is ran in the \emph{logit} domain!

\end{frame}

  \begin{frame}
  \frametitle{The Shapley values in machine learning}

\tikzset{%
  every neuron/.style={
    circle,
    draw,
    minimum size=0.5cm
  },
  neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}

\begin{figure}
  \centering
  \begin{tikzpicture}[>=stealth]

\foreach \m/\l [count=\y] in {1,2,3,missing,4}
  \node [every neuron/.try, neuron \m/.try] (input-\m) at (0,2.5-\y) {};

\foreach \m [count=\y] in {1,missing,2}
  \node [every neuron/.try, neuron \m/.try ] (hidden-\m) at (2,2-\y*1.25) {};

\foreach \m [count=\y] in {1,2,missing,3}
  \node [every neuron/.try, neuron \m/.try ] (output-\m) at (4,2-\y) {};

\foreach \l [count=\i] in {1,2,3,n}
  \draw [<-] (input-\i) -- ++(-1,0)
    node [above, midway] {$x_\l$};

\foreach \l [count=\i] in {1,n}
  \node [above] at (hidden-\i.north) {};

\foreach \l [count=\i] in {1,2,N}
  \draw [->] (output-\i) -- ++(1,0)
  node [above, midway] {$p_\l$};

\foreach \l [count=\i] in {1,2,N}
  \draw [very thick, color=red] (output-\i)+(1,0) circle (0.2);

\foreach \i in {1,...,4}
  \foreach \j in {1,...,2}
    \draw [->] (input-\i) -- (hidden-\j);

\foreach \i in {1,...,2}
  \foreach \j in {1,...,3}
    \draw [->] (hidden-\i) -- (output-\j);

  \end{tikzpicture}
\end{figure}
\begin{center}
  Some explain the output one-by-one,
\end{center}

        \begin{center}
          But a probability distribution lives on a simplex,\\
          \textbf{The relative information matter!!}
        \end{center}

      \end{frame}

  \begin{frame}
  \frametitle{The Shapley values in machine learning}

\tikzset{%
  every neuron/.style={
    circle,
    draw,
    minimum size=0.5cm
  },
  neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}

\begin{figure}
  \centering
  \begin{tikzpicture}[>=stealth]

\foreach \m/\l [count=\y] in {1,2,3,missing,4}
  \node [every neuron/.try, neuron \m/.try] (input-\m) at (0,2.5-\y) {};

\foreach \m [count=\y] in {1,missing,2}
  \node [every neuron/.try, neuron \m/.try ] (hidden-\m) at (2,2-\y*1.25) {};

\foreach \m [count=\y] in {1,2,missing,3}
  \node [every neuron/.try, neuron \m/.try ] (output-\m) at (4,2-\y) {};

\foreach \l [count=\i] in {1,2,3,n}
  \draw [<-] (input-\i) -- ++(-1,0)
    node [above, midway] {$x_\l$};

\foreach \l [count=\i] in {1,n}
  \node [above] at (hidden-\i.north) {};

\foreach \l [count=\i] in {1,2,N}
  \draw [->] (output-\i) -- ++(1,0)
  node [above, midway] {$p_\l$};

\foreach \i in {1,...,4}
  \foreach \j in {1,...,2}
    \draw [->] (input-\i) -- (hidden-\j);

\foreach \i in {1,...,2}
  \foreach \j in {1,...,3}
    \draw [->] (hidden-\i) -- (output-\j);

  \draw [very thick, color=red] (5,-0.5) ellipse (0.2 and 2);

  \end{tikzpicture}
\end{figure}
        \begin{center}
          We will explain the probablities all together using the\\
          \emph{Aitchison geometry of the simplex}.
        \end{center}

  \end{frame}

\section{Compositional data analysis}

\begin{frame}
\frametitle{Compositional data analysis}

\end{frame}

\section{Shapley composition on the simplex}

\begin{frame}
\frametitle{Shapley composition on the simplex}

\end{frame}

\section{Explaining a prediction with Shapley compositions}

\begin{frame}
\frametitle{Explaining a prediction with Shapley compositions}

\end{frame}

\section{Discussion and conclusion}

\begin{frame}
\frametitle{Discussion and conclusion}

\end{frame}

\begin{frame}[t,allowframebreaks]
   \frametitle{References}
    \printbibliography
  \end{frame}

\begin{frame}
\frametitle{Thank you!!}

\end{frame}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2024-01-17 presentation/figures/bifurctree.tex
edit presentation

--- 2023-12-19 paper/bifurctree.tex
@@
-\begin{tikzpicture}[scale=1]
+\begin{tikzpicture}[scale=1.1]
   \draw (0,0) -- (1.1,1.1);
@@
 
-  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{1}$};
-  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{2}$};
-  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{N-1}$};
+  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{\color{blue}$\tilde{p}_{1}$};
+  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{\color{magenta}$\tilde{p}_{2}$};
+  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\color{teal}$\tilde{p}_{N-1}$};
 
   \draw[black] (0,-0.25) node{$p_1$};
+  \draw[white,fill=none](0,-0.25) circle (0.25);
   \draw[black] (1,-0.25) node{$p_2$};

### 2024-01-17 presentation/figures/bifurctree2.tex
edit presentation

--- 2023-12-19 paper/bifurctree.tex
@@
 
-  \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{1}$};
-  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{2}$};
-  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{$\tilde{p}_{N-1}$};
+  \filldraw[black] (0.5,0.5) circle (1.5pt) node[anchor=south east]{\color{blue}\Large$\tilde{p}_{1}$};
+
+  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{\color{red}$\tilde{p}_{2}$};
+  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\color{green}$\tilde{p}_{N-1}$};
 
   \draw[black] (0,-0.25) node{$p_1$};
+  \draw[red,fill=none](0,-0.25) circle (0.25);
   \draw[black] (1,-0.25) node{$p_2$};
+  \draw[red,fill=none](1,-0.25) circle (0.25);
   \draw[black] (2,-0.25) node{$p_3$};

### 2024-01-17 presentation/figures/bifurctree3.tex
edit presentation

--- 2024-01-17 presentation/figures/bifurctree.tex
@@
   \filldraw[black] (0.5,0.5) circle (0.75pt) node[anchor=south east]{\color{blue}$\tilde{p}_{1}$};
-  \filldraw[black] (1,1) circle (0.75pt) node[anchor=south east]{\color{magenta}$\tilde{p}_{2}$};
-  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\color{teal}$\tilde{p}_{N-1}$};
+
+  \filldraw[black] (1,1) circle (1.5pt) node[anchor=south east]{\color{red}\Large$\tilde{p}_{2}$};
+  \filldraw[black] (1.5,1.5) circle (0.75pt) node[anchor=south east]{\color{green}$\tilde{p}_{N-1}$};
 
   \draw[black] (0,-0.25) node{$p_1$};
-  \draw[white,fill=none](0,-0.25) circle (0.25);
   \draw[black] (1,-0.25) node{$p_2$};
+
+  \draw[red,fill=none](.5,-0.25) ellipse (0.75cm  and 0.25cm);
+
   \draw[black] (2,-0.25) node{$p_3$};
+  \draw[red,fill=none](2,-0.25) circle (0.25);
   \draw[black] (3,-0.25) node{$p_{N}$};

### 2024-01-17 presentation/main.tex
edit presentation

@@
 \usepackage{amssymb,amsmath,mathrsfs,stmaryrd,bm,amsthm,amsfonts}
+\usepackage{mathtools}
 \usepackage{pgfplotstable}
 \usepackage{biblatex}
+\usepackage{soul}
 \addbibresource{../paper/biblio.bib}
@@
 \usetikzlibrary{math}
+
+\DeclareMathOperator{\logit}{logit}
+\DeclareMathOperator{\sigmoid}{sigmoid}
+\DeclareMathOperator{\abs}{abs}
+\DeclareMathOperator{\diag}{diag}
+\DeclareMathOperator{\llr}{llr}
+\DeclareMathOperator{\lr}{lr}
+\DeclareMathOperator{\ilr}{ilr}
+\DeclareMathOperator{\ilrl}{ilrl}
+\DeclareMathOperator{\e}{e}
 
@@
     \frametitle{Introduction}
-    \textbf{Shapley values in cooperatibe game theory}\footfullcite{shapley1953value}
+    \textbf{Shapley values in cooperative game theory}\footfullcite{shapley1953value}
     \begin{itemize}
@@
 
-          ex: The output of a softmax layer lives on a multidimensional simplex!
+          ex: The output of a softmax lives on a multidimensional simplex!
         \end{itemize}
@@
 
+      \end{frame}
+
+\begin{frame}{Introduction}
+  \begin{center}
+    The relative information between parts matter!
+  \end{center}
+  \begin{figure}
+    \centering
+\begin{tikzpicture}
+\begin{axis}[
+  ymin=0, ymax=55,
+  yticklabel=\empty,
+  xticklabel=\empty,
+    minor y tick num = 3,
+    area style,
+    ]
+    \addplot+[ybar interval,mark=no] plot coordinates { (0, 0) (5, 31) (10, 0) (15, 0) (20, 0) };
+    \node[] at (axis cs:2.5,15) {\Huge ?};
+    \node[] at (axis cs:12.5,15) {\Huge ?};
+    \node[] at (axis cs:17.5,15) {\Huge ?};
+\end{axis}
+\end{tikzpicture}
+  \end{figure}
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+\begin{frame}{Introduction}
+  \begin{center}
+    The relative information between parts matter!
+  \end{center}
+  \begin{figure}
+    \centering
+\begin{tikzpicture}
+\begin{axis}[
+  ymin=0, ymax=55,
+  yticklabel=\empty,
+  xticklabel=\empty,
+    minor y tick num = 3,
+    area style,
+    ]
+\addplot+[ybar interval,mark=no] plot coordinates { (0, 31) (5, 31) (10, 7) (15, 31) (20, 0) };
+\end{axis}
+\end{tikzpicture}
+  \end{figure}
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+\begin{frame}{Introduction}
+  \begin{center}
+    The relative information between parts matter!
+  \end{center}
+  \begin{figure}
+    \centering
+\begin{tikzpicture}
+\begin{axis}[
+  ymin=0, ymax=55,
+  yticklabel=\empty,
+  xticklabel=\empty,
+    minor y tick num = 3,
+    area style,
+    ]
+\addplot+[ybar interval,mark=no] plot coordinates { (0, 16) (5, 31) (10, 0) (15, 53) (20, 0) };
+\end{axis}
+\end{tikzpicture}
+  \end{figure}
 \end{frame}
@@
   \frametitle{The Shapley values in machine learning}
-The \textbf{value function}:
+The \textbf{prediction difference} (knowing only features indexed by $S$):
 \begin{equation}
   \begin{aligned}
-    v_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
-    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S],
+    \delta_{f,\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathbb{R},\\
+    S &\mapsto \mathbb{E}_\text{Pr}[f(\bm{x})\mid \bm{x}_S] - \mathbb{E}_\text{Pr}[f(\bm{x})],
   \end{aligned}
@@
   \label{eq:contrib}
-  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = v_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - v_{f,\bm{x},\text{Pr}}(\bm{X}_S),
+  c_{f,\bm{x},\text{Pr}}(i,\bm{X}_S) = \delta_{f,\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) - \delta_{f,\bm{x},\text{Pr}}(\bm{X}_S),
 \end{equation}
@@
   \end{figure}
+
+  Efficiency: $\underbrace{f(\bm{x})}_{\text{prediction}}-\underbrace{\mathbb{E}_{\text{Pr}}[f(\bm{X})]}_{\text{base value}} = \sum_{i=1}^{d} \phi_f(i),$
+  \vspace{1cm}
 
@@
 
-\begin{frame}
-\frametitle{Compositional data analysis}
+\begin{frame}{Compositional data analysis}
+  \begin{minipage}{0.49\textwidth}
+    \begin{figure}
+      \centering
+      \includegraphics[width=0.6\textwidth]{figures/basalt.jpg}
+      \caption{A piece of basalt}
+    \end{figure}
+  \end{minipage}
+    \begin{minipage}{0.5\textwidth}
+      Composition:
+      \begin{itemize}
+      \item 35\% of pyroxene,
+      \item 50\% of plagioclase,
+      \item 12\% of olivine,
+      \item 3\% of magnetite.  
+      \end{itemize}
+      \vspace{0.4cm}
+      $\bm{x} = [35,50,12,3]^T$,\\
+      \vspace{0.2cm}
+      $\displaystyle \sum_{i} x_i = 100\%$.
+  \end{minipage}
+\end{frame}
 
+\begin{frame}{Compositional data analysis}
+  A composition lives on a simplex:
+    \vspace{-0.3cm}
+    \begin{equation}
+      \mathcal{S}^N = \Big\{ \bm{x} = [x_1, x_2,\dots x_{N}]^T \in \mathbb{R}^N \mid \forall i\in\llbracket 1,N \rrbracket, x_i>0 \text{~and~} \sum_{i=1}^{N} x_i = k \Big\},
+    \end{equation}
+    \vspace{-0.3cm}
+    \begin{minipage}{0.53\textwidth}
+      \begin{figure}
+        \centering
+        \input{figures/simplex}
+        \caption{Two dimensional simplex as the sample space of three-parts compositional data.}
+      \end{figure}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
+  Exemples of compositional data:
+  \begin{itemize}
+  \item vectors of percentages,
+  \item vectors of concentrations,
+  \item \textbf{discrete probability distributions} (probability simplex: $k=1$).
+    \end{itemize}
+    \end{minipage}    
+\end{frame}
+
+\begin{frame}{Compositional data analysis}
+  \begin{center}
+    {\large
+      \textbf{The Aitchison geometry of the simplex} gives to the simplex an\\
+      \textbf{Euclidean vector space} structure.
+    } \footfullcite{aitchison1986statistical,aitchison2001}\\
+  \end{center}
+  \vspace{0.5cm}
+  {\footnotesize
+    $\bm{x},\bm{y} \in \mathcal{S}^N$ and $\alpha \in \mathbb{R}$,
+  \begin{itemize}
+\item perturbation: $\bm{x}\oplus \bm{y} = \mathcal{C}\left([x_1y_1,\dots x_{N}y_{N}]\right)$,
+    \item powering: $\alpha \odot \bm{x} = \mathcal{C}\left([x_{1}^{\alpha},\dots x_{N}^{\alpha}]\right)$,
+    \item inner product: $\displaystyle \langle \bm{x},\bm{y} \rangle_a = \frac{1}{2N}\sum_{i=1}^{N} \sum_{j=1}^{N} \log \frac{x_i}{x_j}\log \frac{y_i}{y_j}$
+  \end{itemize}
+  where, $\mathcal{C}\left(\bm{x} \right) = \left[ \frac{x_1}{\lVert \bm{x} \rVert_1}, \frac{x_2}{\lVert \bm{x} \rVert_1} ,\dots \frac{x_N}{\lVert \bm{x} \rVert_1} \right]^T$ is the \emph{closure} operator.
+  }
+\end{frame}
+
+\begin{frame}{Compositional data analysis}
+
+  \textbf{Cartesian coordinate system and the ILR transformation}
+\begin{equation}
+    \tilde{\bm{p}}=\ilr(\bm{p}) = \left[\underbrace{\langle \bm{p}, \bm{e}^{(1)} \rangle_a}_{\color{blue} \tilde{\bm{p}}_1} , \underbrace{\langle \bm{p}, \bm{e}^{(2)} \rangle_a}_{\color{magenta} \tilde{\bm{p}}_2},\dots \underbrace{\langle \bm{p}, \bm{e}^{(N-1)} \rangle_a}_{\color{teal} \tilde{\bm{p}}_{N-1}}  \right]^T \in \mathbb{R}^{N-1}.
+  \end{equation}
+  where $\{\bm{e}^{(i)}\}_{1\leq i \leq N}$ forms an \emph{Aitchison} orthonormal basis on the simplex\footfullcite{egozcue2003isometric}.
+  \begin{figure}
+    \begin{minipage}{0.39\textwidth}
+      \centering
+      \input{figures/bifurctree}
+    \end{minipage}
+    \begin{minipage}{0.6\textwidth}
+      \small
+      \begin{equation}
+        \tilde{p}_i = \langle \bm{p}, \bm{e}^{(i)} \rangle_a = \frac{1}{\sqrt{i\,(i+1)}} \log \left( \frac{\prod\limits_{j = 1}^{i}p_j}{(p_{i+1})^{i}} \right).
+      \end{equation}
+    \end{minipage}
+    \caption{Bifurcating tree corresponding to the orthonormal basis obtained with the Gram-Schmidt procedure.}
+    \label{fig:bifurctree}
+  \end{figure}
 \end{frame}
@@
 
+  \begin{center}
+    \large
+    Use the Aitchison geometry to extend the concept of Shapley value to the simplex for explaining multidimensional probabilistic predictions.
+  \end{center}
+  \vspace{1cm}
+\pause
+  We want to explain a prediction $\bm{f}(\bm{x}) \in  \mathcal{S}^N$, where $\bm{f}:\mathcal{X}\to \mathcal{S}^N$ is the learned model.
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+  \node [draw,
+	minimum width=1cm,
+	minimum height=1cm
+        ]  (model) {$\bm{f}$};
+        \node [left=1cm of model] (x) {$\bm{x}$};
+        \node [right=1cm of model] (pred) {\begin{tabular}{c}$\bm{f}(\bm{x})\in \mathcal{S}^N$\\\scriptsize $(N-1)$-dimensional\\\scriptsize probabilistic prediction\end{tabular}};
+        \draw[-stealth] (x.east) -- (model.west);
+        \draw[-stealth] (model.east) -- (pred.west);
+      \end{tikzpicture}
+    \end{figure}
+
+  \end{frame}
+
+  \begin{frame}
+    \frametitle{Shapley composition on the simplex}
+
+    We rewrite the \textbf{prediction difference} and the \textbf{contribution} as follows:
+    \begin{itemize}
+      \pause
+    \item \begin{equation}
+        \begin{aligned}
+          \bm{\delta}_{\bm{f},\bm{x},\text{Pr}}: 2^{\mathcal{I}} &\to \mathcal{S}^N,\\
+          S &\mapsto \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})\mid \bm{x}_S] \ominus \mathbb{E}^{\mathcal{A}}_\text{Pr}[\bm{f}(\bm{x})],
+        \end{aligned}
+      \end{equation}
+      \pause
+      \item \begin{equation}
+        \bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\bm{X}_S) = \bm{\delta}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_{S\cup\{i\}}) \ominus \bm{\delta}_{\bm{f},\bm{x},\text{Pr}}(\bm{X}_S),
+      \end{equation}
+    \end{itemize}
+where $\bm{a}\ominus\bm{b}$ is the perturbation $\bm{a} \oplus \left( (-1)\odot \bm{b}\right)$ which correspond to a substraction between compositions $\bm{a}$ and $\bm{b}$, and where:
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Shapley composition on the simplex}
+  The Shapley quantity expressing the contribution of the $i$th feature on a prediction can simply be expressed on the simplex as the composition $\bm{\phi}(i)$ given by:
+\begin{equation}
+  \bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i) = \frac{1}{d!} \odot \underset{\pi}{\bigoplus}\bm{c}_{\bm{f},\bm{x},\text{Pr}}(i,\pi^{<i}_{\bm{X}}).
+\end{equation}
+We call this quantity {\large\textbf{Shapley composition}}.\\
+It lives on the probability simplex.
+\vspace{0.5cm}
+
+\pause
+It is:
+\begin{itemize}
+\item Linear 
+\item Efficient: $\underset{i=1}{\overset{d}\bigoplus} \bm{\phi}_{\bm{f}}(i) = \bm{f}(\bm{x}) \ominus \mathbb{E}^{\mathcal{A}}_{\text{Pr}}[\bm{f}(\bm{X})]$
+\item Symmetric
+\end{itemize}
+
 \end{frame}
@@
 \frametitle{Explaining a prediction with Shapley compositions}
+\framesubtitle{Visualizations in the isometric-log-ratio space}
 
+We can visualize the Shapley compositions in the isometric-log-ratio space.
+\begin{figure}
+  \centering
+  \input{figures/prob_simplex}
+  \caption{Isometric-log-ratio transformation of the $2$-dimensional simplex (basis obtained with the Gram-Schmidt procedure)}
+\end{figure}
+where $\tilde{p}_1 = \frac{1}{\sqrt{2}} \log \frac{p_1}{p_2}$ and $\tilde{p}_2 = \frac{1}{\sqrt{6}} \log \frac{p_1p_2}{p_3^2}$.
+\end{frame}
+
+\begin{frame}
+\frametitle{Explaining a prediction with Shapley compositions}
+\framesubtitle{Visualizations in the isometric-log-ratio space}
+
+And with more classes?
+\vspace{0.2cm}
+
+We can still visualize subspaces.
+\pause
+\begin{figure}
+  \centering
+  \input{../paper/bifurctree1}
+  \caption{Bifurcation tree used in our $10$-classes digit recognition task.}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Explaining a prediction with Shapley compositions}
+  \framesubtitle{Histograms}
+  \begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{../paper/figures/3classes/histo.png}
+    \caption{Shapley compositions visualized as histograms
+for the Iris classification example}
+  \end{figure}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Explaining a prediction with Shapley compositions}
+  \framesubtitle{Histograms}
+  \begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{../paper/figures/moreclasses/histo.png}
+    \caption{Shapley compositions visualized as histograms for the seven classes digit recognition example.}
+  \end{figure}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Explaining a prediction with Shapley compositions}
+
+  \begin{center}
+    Use the geometrical tools!
+  \end{center}
+  \vspace{0.2cm}
+
+  We can summarize an explanation with:
+  \begin{itemize}
+  \item The norms of the Shapley compositions,
+  \item Angles between them,
+  \item And projection on the class-compositions.
+  \end{itemize}
 \end{frame}
@@
 \frametitle{Discussion and conclusion}
+Summarize:
+\begin{itemize}
+\item We proposed an extention of the concept of Shapley value for explaination in a multiclass setting
+\pause
+\item Shapley composition,
+\pause
+\item \st{scalar} \textbf{distribution/composition} living on the simplex
+  \pause
+\item Visualisations in the ILR space, or as histograms,
+  \pause
+  \item Computation of norms, angles and projections.
+\end{itemize}
 
+\pause
+\vspace{0.2cm}
+Future work:
+\begin{itemize}
+\item The literature about Shapley value is prolific,
+\pause
+\item Axiomatic formulation and uniqueness...
+\end{itemize}
 \end{frame}
@@
 \frametitle{Thank you!!}
+\begin{figure}\centering
+  \includegraphics[width=0.6\linewidth]{bowling.JPG}
+\end{figure}
 
